---
title: "CTCF footprint"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE)
library(tidyverse)
library(patchwork)
library(ggplot2)
library(reshape2)
library(GenomicRanges)
library(rtracklayer)
library(plyranges)
library(vcfR)
library(parallel)
library(Rsamtools)
library(DT)
library(data.table)
library(dplyr)
library(tidyr)
library(purrr)
library(genomation)
library(scales)
library(cowplot)
library(ggpubr)


source("/project/spott/kevinluo/Fiber_seq/fiberhub/code/genotype_utils.R")

# adapted Kevin's code to setup environment
source("/project/spott/cshan/code/process_fiberseq_data.R")
source("/project/spott/kevinluo/Fiber_seq/fiberhub/code/plots.R")

source("/project/spott/kevinluo/Fiber_seq/code/plot_region_functions.R")
source("/project/spott/kevinluo/Fiber_seq/code/fiber_footprint_caller.R")
```



# Kevin's guide to plot specific region of the genome with TF footprints
https://kevinlkx.github.io/fiber-seq-analysis/plot_region_examples.html#Example_region_with_TF_footprints 

```{r}
# Sample table for LCL Fiber-seq data used in QTL analysis
sample_table_file = "/project/spott/kevinluo/Fiber_seq/data/Fiberseq/LCL_sample_metatable_merged_samples_31samples.csv"
sample_table <- data.table::fread(sample_table_file, header = TRUE)

# Select samples of interest
selected_sample_table <- sample_table %>% dplyr::filter(sample_name %in% c("AL48_bc2134_18505"))

# FiberHMM result directory
fiberhmm_result_dir = "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/merged"

# Vierstra motif clustering file
motif_clustering_file = "/project/spott/kevinluo/Fiber_seq/data/motifs/Vierstra_motif_clustering/hg38.archetype_motifs.v1.0.bed.gz"

# FiberHMM MACS peak directory
peak_dir = "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/merged/combined/joint_trained_peaks"
# refined FiberHMM peak directory
refined_peak_dir = "/project/spott/1_Shared_projects/LCL_Fiber_seq/FiberHMM/merged/combined/joint_trained_peaks/refined_peaks"

# Output directory with parsed results for the regions of interest
parsed_result_dir = "/project/spott/cshan/fiber-seq/results/extract_region_result_test"
# create the directory if it does not exist
if (!dir.exists(parsed_result_dir)){
  dir.create(parsed_result_dir, recursive = TRUE)
}

# Set number of cores for parallel processing
ncore = 4
```

```{r}
region_str = "chr5:180328600-180329100"
region_gr <- as(region_str, "GRanges")
region <- as.data.frame(region_gr) %>% dplyr::rename(chr = seqnames) %>% select(chr, start, end)

# select samples
sample_names <- selected_sample_table$sample_name
sample_IDs <- selected_sample_table$cell_line
cat(length(sample_names), "selected samples. \n")
```

```{r}
parsed_res_dir <- file.path(parsed_result_dir, region_str)
result_dirs <- mclapply(1:nrow(selected_sample_table), function(i){
  sample_name = selected_sample_table$sample_name[i]
  sample_ID = selected_sample_table$cell_line[i]
  cat("sample_name:", sample_name, "sample_ID:", sample_ID, "\n")
  bam_file = selected_sample_table$bam_file[i]
  vcf_file = selected_sample_table$phased_vcf_file[i]
  fire_dir = selected_sample_table$fire_dir[i]
  fire_name = selected_sample_table$sample_name[i]
  fiberhmm_file = paste0(fiberhmm_result_dir, "/", sample_name, "/joint_trained_results/", sample_name, "_", region$chr, "_fp.bed.gz")
  if (!file.exists(fiberhmm_file)){
    stop("FiberHMM result does not exist: ", fiberhmm_file, "\n")
  }
  extract_region_result(sample_name, sample_ID, region_str, bam_file, fire_dir, fire_name, 
                        fiberhmm_file, vcf_file, parsed_res_dir, 
                        skip_if_exist = TRUE)
}, mc.cores = ncore)

```

```{r}
result_dirs <- unlist(result_dirs)
names(result_dirs) <- sample_names
selected_sample_table$result_dir <- result_dirs

# print the sample table for selected samples
DT::datatable(selected_sample_table[, c("sample_name", "cell_line", "result_dir")], 
              options = list(pageLength = nrow(selected_sample_table), scrollX = TRUE))

```

Plot region and highlight CTCF motif position chr5:180328835-180328849 (MA0139.2)

```{r}

plot_name = "CTCF"
subtitle = region_str

# FP peaks to show in the plot
peak_file_10_30bp = file.path(refined_peak_dir, "combined_all_chrs_10-30bp_macs3_q0.01_minratio0.5_refined_peaks.bed.gz")
peak_file_30_60bp = file.path(refined_peak_dir, "combined_all_chrs_30-60bp_macs3_q0.01_minratio0.5_refined_peaks.bed.gz")
peak_file_60_80bp = file.path(refined_peak_dir, "combined_all_chrs_60-80bp_macs3_q0.01_minratio0.5_refined_peaks.bed.gz")
peaks_10_30bp <- read_peaks_region(peak_file_10_30bp, region)
peaks_30_60bp <- read_peaks_region(peak_file_30_60bp, region)
peaks_60_80bp <- read_peaks_region(peak_file_60_80bp, region)

FP_peaks_list <- list(
  "10-30bp FP" = peaks_10_30bp,
  "30-60bp FP" = peaks_30_60bp,
  "60-80bp FP" = peaks_60_80bp)

# Top motif clusters in the motif region
# chr5:180328835-180328849 (MA0139.2)
motif_region <- data.frame(chr = "chr5", start = 180328835, end = 180328849)

top_motif_clusters_list <- read_top_motif_clusters_region(motif_clustering_file, motif_region, 
                                                          minoverlap = 1, min_score = 5, top_n = 5, 
                                                          verbose = TRUE)


```

load CFCF ChIP-seq data
```{r}
extraCols_narrowPeak <- c(signalValue = "numeric", pValue = "numeric",
                          qValue = "numeric", peak = "integer")


ctcf_peaks <- rtracklayer::import(
  "/project/spott/cshan/annotations/CTCF_ENCFF960ZGP_narrowPeak.bed",
  format = "BED",
  extraCols = extraCols_narrowPeak
) %>%
  as.data.frame() %>%
  dplyr::rename(chr = seqnames)

ctcf_peaks$chr <- as.character(ctcf_peaks$chr)

# subset CTCF peaks to region
ctcf_peaks_region <- ctcf_peaks %>%
  dplyr::filter(chr == region$chr,
                end >= region$start,
                start <= region$end)

ctcf_peaks_region <- ctcf_peaks_region %>%
  dplyr::select(chr, start, end)

nrow(ctcf_peaks_region)



# CTCF bigwig

ctcf_bw <- rtracklayer::import(
  "/project/spott/cshan/annotations/CTCF_ENCFF636ODI.bigWig",
  which = GRanges(region$chr, IRanges(region$start, region$end))) %>%
  as.data.frame()

ctcf_bw <- ctcf_bw %>%
  dplyr::rename(chr = seqnames) %>%
  dplyr::mutate(chr = as.character(chr))

ctcf_track <- ggplot(ctcf_bw) +
  geom_area(aes(x = start, y = score), fill = "steelblue", alpha = 0.7) +
  xlim(region$start, region$end) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_blank()
  )

print(ctcf_track)
```

Overlay CTCF peaks with DNase-seq

DNase-seq data is obtained from: https://www.encodeproject.org/experiments/ENCSR000EMT/ 
  - https://www.encodeproject.org/files/ENCFF743ULW/@@download/ENCFF743ULW.bigWig
  - https://www.encodeproject.org/files/ENCFF073ORT/@@download/ENCFF073ORT.bed.gz 
  
  

```{r}
# reformat Dnase bed file to add back column names

extraCols_narrowPeak <- c(signalValue = "numeric", pValue = "numeric",
                          qValue = "numeric", peak = "integer")

dnase_peaks <- rtracklayer::import(
  "/project/spott/cshan/annotations/DNase_ENCFF073ORT.bed", format = "BED",extraCols = extraCols_narrowPeak) %>%
  as.data.frame() %>%
  dplyr::rename(chr = seqnames)

dnase_peaks$chr <- as.character(dnase_peaks$chr)

# subset CTCF peaks to region
dnase_peaks_region <- dnase_peaks %>%
  dplyr::filter(chr == region$chr,
                end >= region$start,
                start <= region$end)

dnase_peaks_region <- dnase_peaks_region %>%
  dplyr::select(chr, start, end)

nrow(dnase_peaks_region)



# Dnase bigwig
dnase_bw <- rtracklayer::import(
  "/project/spott/cshan/annotations/DNase.ENCFF743ULW.bigWig",
  which = GRanges(region$chr, IRanges(region$start, region$end))) %>%
  as.data.frame()

dnase_bw <- dnase_bw %>%
  dplyr::rename(chr = seqnames) %>%
  dplyr::mutate(chr = as.character(chr))

dnase_track <- ggplot(dnase_bw) +
  geom_area(aes(x = start, y = score), fill = "darkred", alpha = 0.7) +
  xlim(region$start, region$end) +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_blank()
  )

print(dnase_track)

```


```{r}
combined_peaks_df$group <- factor(combined_peaks_df$group, 
                                   levels = rev(c("CTCF ChIP-seq", "DNase-seq")))
combined_peaks_df$y <- as.numeric(combined_peaks_df$group)

combined_peak_track <- ggplot(combined_peaks_df, 
                               aes(x = start, xend = end, y = y, yend = y, color = group)) +
  geom_segment(linewidth = 1) +
  scale_color_manual(values = c("CTCF ChIP-seq" = "steelblue", 
                                 "DNase-seq" = "darkred")) +
  labs(x = NULL, y = NULL) +
  scale_x_continuous(limits = c(region$start, region$end), expand = c(0, 0)) +
  scale_y_continuous(breaks = sort(unique(combined_peaks_df$y)),
                     labels = levels(combined_peaks_df$group),
                     expand = expansion(add = 0.5)) +
  theme(legend.position = "none") +
  coord_cartesian(xlim = c(region$start, region$end), expand = TRUE)

print(combined_peak_track)
```

```{r}
peaks_list <- c(
  FP_peaks_list,
  top_motif_clusters_list)

# Cluster FiberHMM short footprints in the regions
cluster_regions <- motif_region
cluster_regions$start <- motif_region$start - 10
cluster_regions$end <- motif_region$end + 10

# combine results in the region and make plots
res <- get_region_combined_results(result_dirs = result_dirs,
                                   region = region,
                                   sample_names = sample_names,
                                   sample_IDs = sample_IDs,
                                   peaks = peaks_list,
                                   select_A_only = TRUE,
                                   fiberHMM_size_breaks = c(30, 60, 80),
                                   fiberHMM_cols = c("purple", "magenta", "#41c4e1", "darkgreen"),
                                   cover_region_only = TRUE,
                                   cover_cluster_regions_only = TRUE,
                                   remove_unknown_tag = TRUE,
                                   remove_unphased = TRUE,
                                   cluster = "fiberHMM_configs",
                                   cluster_regions = cluster_regions,
                                   cluster_fp_size = c(10, 30),
                                   config_levels = c("0", "1"),
                                   highlight_window = motif_region,
                                   color_highlight_window = "#50d844",
                                   linewidth = 0.5,
                                   smooth_pileup = TRUE,
                                   window_n = 10,
                                   plot_name = plot_name,
                                   subtitle = subtitle,
                                   verbose = TRUE)


```


```{r}
gg <- res$gg

# Combine all plots
gg_combined <- gg$pileup_groups +
  gg$peaks + ylab("") +
  gg$fire_fdr + ylab("") +
  combined_peak_track +   # Combined CTCF and DNase peaks track
  ctcf_track +  
  dnase_track +
  (gg$dimelo_reads + ylab("SMF reads")) +
  (gg$fire_fiberHMM + xlab(region$chr) + ylab("FIRE polished FiberHMM") +
     theme(legend.position = 'right')) + 
  plot_layout(ncol = 1, heights = c(4, 3, 1, 2, 2, 2, 8, 8))

print(gg_combined)

# Save the plot
pdf("/project/spott/cshan/fiber-seq/results/extract_region_result_test/ctcf_chr5_180328600_180329100.pdf", 
    height = 11, width = 12)
gg_combined
dev.off()


```


# Find CTCF motifs across the genome

I want to first prepare the reference genome for TF and interest, as weell as the motifs for TF of interest


- TSS from GENCODE: GetTsite was used to extract TSS from gencode GTF file and export as bed format (https://github.com/junjunlab/GetTsite)
- CTCF ChIP-seq in LCL (GM12878): https://www.encodeproject.org/experiments/ENCSR000DZN/

```{r}
example_ctcf_fp <- fread("/project/spott/cshan/fiber-seq/results/per_sample/AL10_bc2178_19130.5mC.6mA.nuc.aligned/footprints/CTCF_footprint_ge20fibers.clean.tsv")

example_ctcf_fp
```

I ran this script to get the CTCF footprint from motifs /project/spott/cshan/fiber-seq/code/ctcf_footprinting_slurm.sh


```{r ctcf-tf, warning=FALSE}
#===============================================================================
# CTCF Footprint Post-Processing in R
#
# 1. Parse footprint codes from ft footprint output 
# 2. Summarize across all 34 samples
# 3. Generate plots

#===============================================================================



#===============================================================================
# PATHS
#===============================================================================
results_dir <- "/project/spott/cshan/fiber-seq/results"
plot_dir    <- file.path(results_dir, "plots")
dir.create(plot_dir, showWarnings = FALSE, recursive = TRUE)

Sys.setenv(TMPDIR = "/project/spott/cshan/fiber-seq/results/tmp")
dir.create(Sys.getenv("TMPDIR"), showWarnings = FALSE, recursive = TRUE)

#===============================================================================
# STEP 1: Parse footprint codes for all samples
#===============================================================================
cat("=== STEP 1: Parse footprint codes ===\n")

# Find all filtered footprint files
fp_files <- list.files(
  file.path(results_dir, "per_sample"),
  pattern = "CTCF_footprint_ge20fibers.tsv",
  recursive = TRUE,
  full.names = TRUE
)

cat(sprintf("Found %d samples\n", length(fp_files)))

# Function to parse footprint codes for one sample
parse_footprint_codes <- function(fp_file) {
  
  sample_name <- basename(dirname(dirname(fp_file)))
  cat(sprintf("  Parsing: %s\n", sample_name))
  
  dt <- fread(fp_file, sep = "\t")
  
  # Fix column names
  setnames(dt, "#chrom", "chrom")
  setnames(dt, c("start", "end"), c("fp_start", "fp_end"), skip_absent = TRUE)
  
  dt[, site_id := paste(chrom, fp_start, fp_end, strand, sep = ":")]
  
  result <- dt[, {
    codes <- as.integer(strsplit(as.character(footprint_codes), ",")[[1]])
    n <- length(codes)
    
    has_msp <- bitwAnd(codes, 1L) > 0L
    has_fp  <- bitwAnd(codes, 2L) > 0L
    
    n_inacc   <- sum(!has_msp)
    n_unbound <- sum(has_msp & !has_fp)
    n_bound   <- sum(has_msp & has_fp)
    
    list(
      n_fibers              = n,
      n_bound               = n_bound,
      n_accessible_unbound  = n_unbound,
      n_inaccessible        = n_inacc,
      pct_bound             = n_bound / n * 100,
      pct_accessible_unbound = n_unbound / n * 100,
      pct_inaccessible      = n_inacc / n * 100
    )
  }, by = .(chrom, fp_start, fp_end, strand, site_id)]
  
  result[, sample := sample_name]
  setnames(result, c("fp_start", "fp_end"), c("start", "end"))
  
  cat(sprintf("    %d sites, %d total fibers\n", nrow(result), sum(result$n_fibers)))
  cat(sprintf("    Bound: %.1f%% | Unbound: %.1f%% | Inaccessible: %.1f%%\n",
              sum(result$n_bound) / sum(result$n_fibers) * 100,
              sum(result$n_accessible_unbound) / sum(result$n_fibers) * 100,
              sum(result$n_inaccessible) / sum(result$n_fibers) * 100))
  
  return(result)
}

# Parse all samples
all_parsed <- map_dfr(fp_files, parse_footprint_codes)

# Filter out AL29 samples
all_parsed <- all_parsed %>%
  filter(!grepl("^AL29", sample))

# Save combined per-site data
fwrite(all_parsed, file.path(results_dir, "CTCF_all_samples_per_site.tsv"), sep = "\t")
cat(sprintf("\nSaved: CTCF_all_samples_per_site.tsv (%d rows)\n", nrow(all_parsed)))

#===============================================================================
# STEP 2: Cross-sample summary
#===============================================================================
dir.create(file.path(results_dir, "tmp"), showWarnings = FALSE)
Sys.setenv(TMPDIR = file.path(results_dir, "tmp"))

cat("\n=== STEP 2: Cross-sample summary ===\n")

# Extract short sample name (AL## part)
sample_summary <- all_parsed %>%
  group_by(sample) %>%
  summarise(
    n_sites       = n(),
    total_fibers  = sum(n_fibers),
    total_bound   = sum(n_bound),
    total_unbound = sum(n_accessible_unbound),
    total_inacc   = sum(n_inaccessible),
    pct_bound     = total_bound / total_fibers * 100,
    pct_unbound   = total_unbound / total_fibers * 100,
    pct_inacc     = total_inacc / total_fibers * 100,
    mean_site_pct_bound   = mean(pct_bound),
    mean_site_pct_unbound = mean(pct_accessible_unbound),
    mean_site_pct_inacc   = mean(pct_inaccessible),
    .groups = "drop"
  ) %>%
  mutate(short_name = stringr::str_extract(sample, "AL\\d+"))

fwrite(sample_summary, file.path(results_dir, "CTCF_footprint_summary_all_samples.tsv"), sep = "\t")
cat(sprintf("Saved summary for %d samples\n", nrow(sample_summary)))
print(sample_summary %>% select(short_name, n_sites, total_fibers, pct_bound, pct_unbound, pct_inacc))
```


