---
title: "Fastopics_test"
author: "Crystal Shan"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = FALSE
)

library(data.table)
library(dplyr)
library(tidyr)
library(Matrix)
library(GenomicRanges)
library(IRanges)
library(GenomeInfoDb)
library(BiocParallel)
library(ggplot2)
library(fastTopics)

source("/project/spott/cshan/fiber-seq/code/footprintR_modbam_code/modbam_footprintR_functions.R")
```

# Goal

Pilot whether `fastTopics` can learn latent fiber-level methylation patterns from a binary methylation matrix at a subset of transcription-factor binding sites.

For this first pass, treat each fiber/read like a "cell" and each TF-site feature like a "gene":

- rows: fiber reads
- columns: TF binding-site features, starting with KLF1/KLF-family motifs
- values: `1` if the read has a modified 5mC call in the selected TF-site window, otherwise `0`

# Run Environment

Run the notebook from an interactive R session after loading R:

```{bash load-r, eval=FALSE}
module load R
Rscript -e "cat(R.version.string, '\n')"
```



# Parameters

```{r parameters}
sample_name <- Sys.getenv("SAMPLE", unset = "AL10_bc2178_19130")
chrom <- Sys.getenv("CHROM", unset = "chr19")

tf_family <- Sys.getenv("TF_FAMILY", unset = "KLF")
tf_names <- strsplit(Sys.getenv("TF_NAMES", unset = "KLF1"), ",")[[1]]
tf_names <- toupper(trimws(tf_names))

modbase_code <- Sys.getenv("MODBASE_CODE", unset = "m") # "m" = 5mC, "a" = m6A
mod_prob_threshold <- as.numeric(Sys.getenv("MOD_PROB_THRESHOLD", unset = "0.9"))

motif_flank_bp <- as.integer(Sys.getenv("MOTIF_FLANK_BP", unset = "100"))
max_sites_per_tf <- as.integer(Sys.getenv("MAX_SITES_PER_TF", unset = "250"))
min_reads_per_site <- as.integer(Sys.getenv("MIN_READS_PER_SITE", unset = "10"))
min_sites_per_read <- as.integer(Sys.getenv("MIN_SITES_PER_READ", unset = "2"))
k_grid <- as.integer(strsplit(Sys.getenv("K_GRID", unset = "2,3,4,5,6"), ",")[[1]])

project_root <- "/project/spott/cshan/fiber-seq"
homer_motif_dir <- "/project/spott/cshan/tools/homer/motifs"
fiberhub_script_dir <- "/project/spott/kevinluo/Fiber_seq/fiberhub/scripts"
jaspar_motif_dir <- "/project/spott/kevinluo/Fiber_seq/data/motifs/JASPAR2024_CORE_non-redundant_pfms_meme"
sequence_file <- "/project/spott/kevinluo/Fiber_seq/data/ref_genome/hg38.fa"
blacklist_file <- "/project/spott/kevinluo/Fiber_seq/data/ENCODE_blacklist/blacklist.hg38.bed.gz"
chromsize_file <- "/project/spott/kevinluo/Fiber_seq/data/ref_genome/hg38.chrom.sizes"
out_dir <- file.path(project_root, "results", "fastTopics_test", sample_name, paste(tf_names, collapse = "_"))
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

bam_path <- file.path(
  "/project/spott/1_Shared_projects/LCL_Fiber_seq/preprocess_final_merged_samples",
  sprintf("%s.5mC.6mA.aligned.phased.bam", sample_name)
)

motif_sites_tsv <- file.path(out_dir, sprintf("%s.%s.motif_sites.tsv.gz", chrom, paste(tf_names, collapse = "_")))
matrix_rds <- file.path(out_dir, sprintf("%s.%s.binary_methylation_matrix.rds", chrom, paste(tf_names, collapse = "_")))
fit_rds <- file.path(out_dir, sprintf("%s.%s.fastTopics_fits.rds", chrom, paste(tf_names, collapse = "_")))

params <- list(
  sample_name = sample_name,
  chrom = chrom,
  tf_family = tf_family,
  tf_names = tf_names,
  modbase_code = modbase_code,
  mod_prob_threshold = mod_prob_threshold,
  motif_flank_bp = motif_flank_bp,
  max_sites_per_tf = max_sites_per_tf,
  min_reads_per_site = min_reads_per_site,
  min_sites_per_read = min_sites_per_read,
  k_grid = k_grid,
  bam_path = bam_path,
  jaspar_motif_dir = jaspar_motif_dir,
  sequence_file = sequence_file,
  blacklist_file = blacklist_file,
  out_dir = out_dir
)
params
```

# Step 1: Scan Motifs for Specific TFs with TOP/FIMO

Start with KLF binding sites, especially KLF1. This section borrows the motif-scanning pattern from:

- `/project/spott/kevinluo/Fiber_seq/fiberhub/scripts/get_motif_sites.R`
- `/project/spott/kevinluo/Fiber_seq/fiberhub/scripts/motifbreak_analysis_ASfp_jaspar_motifs.R`

The key pieces to reuse are:

- `TOP::fimo_motif_matches()` to scan hg38 with FIMO.
- `TOP::process_candidate_sites()` to apply p-value/PWM cutoffs, add flanks, remove ENCODE blacklist overlaps, and return clean candidate sites.
- JASPAR MEME motif files from `/project/spott/kevinluo/Fiber_seq/data/motifs/JASPAR2024_CORE_non-redundant_pfms_meme`.

For KLF-family motifs in this JASPAR2024 directory:

- KLF1: `MA0493.3`
- KLF3: `MA1516.2`
- KLF4: `MA0039.5`
- KLF5: `MA0599.1`
- KLF6: `MA1517.2`
- KLF9: `MA1107.3`
- KLF10: `MA1511.2`
- KLF14: `MA0740.2`
- KLF15: `MA1513.2`
- KLF17: `MA1514.2`

```{r motif-scan-params}
thresh_pvalue_label <- Sys.getenv("MOTIF_THRESH_P", unset = "1e-5")
thresh_pvalue <- as.numeric(thresh_pvalue_label)
thresh_pwm_score <- as.numeric(Sys.getenv("MOTIF_THRESH_PWM", unset = "10"))

tf_to_jaspar_motif <- function(tf) {
  tf_upper <- toupper(tf)

  motif_id <- dplyr::case_when(
    tf_upper == "KLF1" ~ "MA0493.3",
    tf_upper == "KLF3" ~ "MA1516.2",
    tf_upper == "KLF4" ~ "MA0039.5",
    tf_upper == "KLF5" ~ "MA0599.1",
    tf_upper == "KLF6" ~ "MA1517.2",
    tf_upper == "KLF9" ~ "MA1107.3",
    tf_upper == "KLF10" ~ "MA1511.2",
    tf_upper == "KLF14" ~ "MA0740.2",
    tf_upper == "KLF15" ~ "MA1513.2",
    tf_upper == "KLF17" ~ "MA1514.2",
    TRUE ~ NA_character_
  )

  data.table(
    tf = tf_upper,
    motif_id = motif_id,
    motif_file = file.path(jaspar_motif_dir, paste0(motif_id, ".meme")),
    motif_exists = !is.na(motif_id) && file.exists(file.path(jaspar_motif_dir, paste0(motif_id, ".meme")))
  )
}

tf_motif_table <- data.table::rbindlist(lapply(tf_names, tf_to_jaspar_motif))
tf_motif_table
```

## Option A: Call Kevin's motif-site script

This calls the existing script directly once per TF. It writes FIMO output, candidate-site RDS, and candidate-site BED files.

```{bash call-kevin-get-motif-sites, eval=FALSE}
module load R

SAMPLE=${SAMPLE:-AL10_bc2178_19130}
TF_NAMES=${TF_NAMES:-KLF1}
OUT_DIR=/project/spott/cshan/fiber-seq/results/fastTopics_test/${SAMPLE}/${TF_NAMES//,/_}/motif_sites
mkdir -p "${OUT_DIR}"

IFS=',' read -ra TFS <<< "${TF_NAMES}"
for TF in "${TFS[@]}"; do
  TF_UPPER=$(echo "${TF}" | tr '[:lower:]' '[:upper:]')

  case "${TF_UPPER}" in
    KLF1) MOTIF_ID="MA0493.3" ;;
    KLF3) MOTIF_ID="MA1516.2" ;;
    KLF4) MOTIF_ID="MA0039.5" ;;
    KLF5) MOTIF_ID="MA0599.1" ;;
    KLF6) MOTIF_ID="MA1517.2" ;;
    KLF9) MOTIF_ID="MA1107.3" ;;
    KLF10) MOTIF_ID="MA1511.2" ;;
    KLF14) MOTIF_ID="MA0740.2" ;;
    KLF15) MOTIF_ID="MA1513.2" ;;
    KLF17) MOTIF_ID="MA1514.2" ;;
    *) echo "Add JASPAR motif ID for ${TF_UPPER}" >&2; exit 1 ;;
  esac

  Rscript /project/spott/kevinluo/Fiber_seq/fiberhub/scripts/get_motif_sites.R \
    --tf "${TF_UPPER}" \
    --motif "${MOTIF_ID}" \
    --threshP 1e-5 \
    --threshPWM 10 \
    --flank 100 \
    --outdir "${OUT_DIR}"
done
```

## Option B: Run the borrowed TOP/FIMO logic in this notebook

This is the same idea as `get_motif_sites.R`, but keeps the outputs under this notebook's `out_dir` and adds a `tf` column before merging TFs.

```{r scan-tf-motifs-with-top, eval=FALSE}
library(TOP)

motif_out_dir <- file.path(out_dir, "motif_sites")
dir.create(motif_out_dir, recursive = TRUE, showWarnings = FALSE)

if (!file.exists(chromsize_file)) {
  TOP::index_fa(sequence_file, chromsize_file = chromsize_file)
}

scan_one_tf <- function(tf, motif_id) {
  motif_file <- file.path(jaspar_motif_dir, paste0(motif_id, ".meme"))
  assert_file_exists(motif_file, paste("JASPAR motif", motif_id))

  fimo_file <- file.path(motif_out_dir, sprintf("%s_%s_%s.fimo.txt", tf, motif_id, thresh_pvalue_label))
  candidate_rds <- file.path(motif_out_dir, sprintf("%s_%s_%s.candidate.sites.rds", tf, motif_id, thresh_pvalue_label))
  candidate_bed <- file.path(motif_out_dir, sprintf("%s_%s_%s.candidate.sites.bed", tf, motif_id, thresh_pvalue_label))

  TOP::fimo_motif_matches(
    motif_file = motif_file,
    sequence_file = sequence_file,
    thresh_pValue = thresh_pvalue,
    outname = fimo_file,
    fimo_path = "fimo"
  )

  sites <- TOP::process_candidate_sites(
    fimo_file = fimo_file,
    flank = motif_flank_bp,
    thresh_pValue = thresh_pvalue,
    thresh_pwmscore = thresh_pwm_score,
    blacklist_file = blacklist_file,
    chr_order = paste0("chr", 1:22)
  )

  sites <- data.table::as.data.table(sites)
  sites[, tf := tf]
  sites[, motif_id := motif_id]

  saveRDS(sites, candidate_rds)
  data.table::fwrite(sites[, 1:6], candidate_bed, sep = "\t", col.names = FALSE)
  sites
}

stopifnot(all(tf_motif_table$motif_exists))

motif_sites_all <- data.table::rbindlist(
  lapply(seq_len(nrow(tf_motif_table)), function(i) {
    scan_one_tf(tf_motif_table$tf[i], tf_motif_table$motif_id[i])
  }),
  use.names = TRUE,
  fill = TRUE
)

data.table::fwrite(motif_sites_all, motif_sites_tsv, sep = "\t")
```


```{r read-motif-sites, eval=FALSE}
standardize_top_sites <- function(path, chrom, max_sites_per_tf = 250L) {
  x <- data.table::fread(path)

  if ("chr" %in% names(x) && !"chromosome" %in% names(x)) {
    data.table::setnames(x, "chr", "chromosome")
  }
  if ("start" %in% names(x) && !"start0" %in% names(x)) {
    data.table::setnames(x, "start", "start0")
  }
  if ("end" %in% names(x) && !"end0" %in% names(x)) {
    data.table::setnames(x, "end", "end0")
  }

  required <- c("chromosome", "start0", "end0", "tf", "motif_id")
  missing <- setdiff(required, names(x))
  if (length(missing)) {
    stop("Missing required motif-site columns: ", paste(missing, collapse = ", "), call. = FALSE)
  }

  if (!"strand" %in% names(x)) {
    x[, strand := "*"]
  }
  if (!"score" %in% names(x)) {
    score_col <- intersect(c("pwm_score", "score_fwd", "score_rev"), names(x))[1]
    x[, score := if (!is.na(score_col)) get(score_col) else NA_real_]
  }

  x[, `:=`(
    chromosome = as.character(chromosome),
    start0 = as.integer(start0),
    end0 = as.integer(end0),
    tf = toupper(as.character(tf)),
    motif_id = as.character(motif_id),
    strand = as.character(strand),
    score = suppressWarnings(as.numeric(score))
  )]

  x <- x[chromosome == chrom]
  x <- x[order(tf, -score)]
  x <- x[, head(.SD, max_sites_per_tf), by = tf]
  x[, site_id := sprintf("%s_%s_%s_%d_%d_%06d", tf, motif_id, chromosome, start0, end0, seq_len(.N)), by = .(tf, motif_id)]
  x[]
}

motif_sites <- standardize_top_sites(
  motif_sites_tsv,
  chrom = chrom,
  max_sites_per_tf = max_sites_per_tf
)

data.table::fwrite(motif_sites, motif_sites_tsv, sep = "\t")
motif_sites
```

# Step 2: Optional LCL-Relevance Filter

After KLF1 works, rank or filter TFs/sites using LCL resources:

- high-expression TFs: TF gene has LCL CAGE/expression support
- active nearby sites: motif site overlaps or is near FANTOM5 LCL CAGE peaks/cCREs
- low-expression control TFs: same motif-scan procedure, but TF gene has little/no LCL evidence

Useful local resources:

- `/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.withGene.bed.gz`
- `/project/spott/cshan/annotations/ENCFF715BEB_CAGE_withGenes.bed`
- `/project/spott/cshan/annotations/JASPAR2026_CORE_non-redundant_pfms_meme.txt`

```{r lcl-filter-placeholder, eval=FALSE}
cage_path <- "/project/spott/cshan/annotations/fantom5/fantom5.hg38.LCL.consensus.CAGE_peaks.withGene.bed.gz"
cage <- data.table::fread(cage_path, header = FALSE)

# TODO: standardize CAGE columns, convert motif_sites and CAGE peaks to GRanges,
# then annotate motif_sites with distance to nearest LCL CAGE peak or overlap.
```

# Step 3: Make TF-Site Windows

```{r motif-windows, eval=FALSE}
motif_sites <- data.table::fread(motif_sites_tsv)
bam_seqinfo <- read_bam_seqinfo(bam_path)

motif_gr <- GRanges(
  seqnames = motif_sites$chromosome,
  ranges = IRanges(start = motif_sites$start0 + 1L, end = motif_sites$end0),
  strand = motif_sites$strand
)

GenomeInfoDb::seqinfo(motif_gr) <- bam_seqinfo[unique(as.character(seqnames(motif_gr)))]

motif_mid <- round((start(motif_gr) + end(motif_gr)) / 2)
motif_windows <- GRanges(
  seqnames = seqnames(motif_gr),
  ranges = IRanges(
    start = pmax(1L, motif_mid - motif_flank_bp),
    end = motif_mid + motif_flank_bp
  ),
  strand = strand(motif_gr)
)
GenomeInfoDb::seqinfo(motif_windows) <- GenomeInfoDb::seqinfo(motif_gr)
motif_windows <- GenomicRanges::trim(motif_windows)

mcols(motif_windows)$site_id <- motif_sites$site_id
mcols(motif_windows)$tf <- motif_sites$tf
mcols(motif_windows)$motif_score <- motif_sites$score

motif_meta <- data.table(
  site_id = motif_sites$site_id,
  tf = motif_sites$tf,
  chromosome = as.character(seqnames(motif_windows)),
  window_start = start(motif_windows),
  window_end = end(motif_windows),
  motif_start0 = motif_sites$start0,
  motif_end0 = motif_sites$end0,
  motif_score = motif_sites$score
)

data.table::fwrite(
  motif_meta,
  file.path(out_dir, sprintf("%s.%s.motif_windows.tsv", chrom, paste(tf_names, collapse = "_"))),
  sep = "\t"
)
motif_meta
```

# Step 4: Read ModBAM and Binarize Methylation

This uses footprintR at read level, then thresholds modification probability into binary calls. For CpG methylation, keep `modbase_code = "m"`.

```{r read-level-modbam, eval=FALSE}
assert_file_exists(bam_path, "BAM")
assert_file_exists(paste0(bam_path, ".bai"), "BAM index")

bamfiles <- setNames(bam_path, sample_name)

se_tf <- footprintR::readModBam(
  bamfiles = bamfiles,
  regions = motif_windows,
  modbase = setNames(modbase_code, sample_name),
  level = "quickread",
  seqinfo = bam_seqinfo,
  trim = TRUE,
  BPPARAM = BiocParallel::SerialParam(),
  verbose = TRUE
)

se_tf <- footprintR::flattenReadLevelAssay(se_tf)

binary_calls <- getBinarizedModificationsLongTable(
  footprintRObject = se_tf,
  sampleName = sample_name,
  modProbThreshold = mod_prob_threshold
)

binary_calls <- data.table::as.data.table(binary_calls)

data.table::fwrite(
  binary_calls,
  file.path(out_dir, sprintf("%s.%s.read_level_binary_calls.tsv.gz", chrom, paste(tf_names, collapse = "_"))),
  sep = "\t"
)
```

```{r assign-calls-to-sites, eval=FALSE}
binary_calls_path <- file.path(out_dir, sprintf("%s.%s.read_level_binary_calls.tsv.gz", chrom, paste(tf_names, collapse = "_")))
binary_calls <- data.table::fread(binary_calls_path)

binary_calls_gr <- GRanges(
  seqnames = binary_calls$call_chromosome,
  ranges = IRanges(start = binary_calls$position, end = binary_calls$position),
  strand = "*"
)

hits <- GenomicRanges::findOverlaps(binary_calls_gr, motif_windows, ignore.strand = TRUE)

calls_by_site <- cbind(
  binary_calls[S4Vectors::queryHits(hits)],
  motif_meta[S4Vectors::subjectHits(hits), .(site_id, tf, chromosome, window_start, window_end)]
)

calls_by_site[, rel_pos := position - round((window_start + window_end) / 2)]

data.table::fwrite(
  calls_by_site,
  file.path(out_dir, sprintf("%s.%s.binary_calls_by_site.tsv.gz", chrom, paste(tf_names, collapse = "_"))),
  sep = "\t"
)
```

# Step 5: Build Binary Matrix for fastTopics

For the pilot, collapse each read-site pair to one value:

- `1`: at least one modified 5mC call in the motif-centered window
- `0`: no modified call observed in that read-site feature

Important caveat: in this simple skeleton, zeros mix true unmodified covered sites and missing coverage. Keep the coverage QC tables so the next iteration can mask missing read-site pairs more explicitly.

```{r build-site-matrix, eval=FALSE}
calls_by_site <- data.table::fread(
  file.path(out_dir, sprintf("%s.%s.binary_calls_by_site.tsv.gz", chrom, paste(tf_names, collapse = "_")))
)

site_binary <- calls_by_site[, .(
  methylated = as.integer(any(mod_status == 1L)),
  covered_calls = .N,
  modified_calls = sum(mod_status == 1L)
), by = .(read_id, site_id)]

site_qc <- site_binary[, .(
  n_reads = .N,
  n_methylated_reads = sum(methylated),
  methylated_fraction = mean(methylated)
), by = site_id]

read_qc <- site_binary[, .(
  n_sites = .N,
  n_methylated_sites = sum(methylated)
), by = read_id]

keep_sites <- site_qc[n_reads >= min_reads_per_site, site_id]
keep_reads <- read_qc[n_sites >= min_sites_per_read, read_id]

site_binary <- site_binary[site_id %in% keep_sites & read_id %in% keep_reads]

read_index <- data.table(read_id = sort(unique(site_binary$read_id)))
read_index[, row_id := .I]

site_index <- data.table(site_id = sort(unique(site_binary$site_id)))
site_index[, col_id := .I]

matrix_dt <- merge(site_binary, read_index, by = "read_id")
matrix_dt <- merge(matrix_dt, site_index, by = "site_id")
matrix_dt <- matrix_dt[methylated == 1L]

X <- Matrix::sparseMatrix(
  i = matrix_dt$row_id,
  j = matrix_dt$col_id,
  x = 1L,
  dims = c(nrow(read_index), nrow(site_index)),
  dimnames = list(read_index$read_id, site_index$site_id)
)

matrix_bundle <- list(
  X = X,
  read_index = read_index,
  site_index = site_index,
  site_qc = site_qc,
  read_qc = read_qc,
  motif_meta = motif_meta,
  params = params
)

saveRDS(matrix_bundle, matrix_rds)
matrix_bundle
```

# Step 6: Fit fastTopics

`fastTopics::fit_topic_model()` expects a non-negative count matrix; this pilot gives it a sparse binary count matrix.

```{r fit-fasttopics, eval=FALSE}
library(fastTopics)

matrix_bundle <- readRDS(matrix_rds)
X <- matrix_bundle$X

stopifnot(nrow(X) > 1L, ncol(X) > 1L)
stopifnot(all(X@x >= 0))

fits <- lapply(k_grid, function(k) {
  set.seed(1)
  fastTopics::fit_topic_model(
    X,
    k = k,
    numiter.main = 100,
    numiter.refine = 100,
    verbose = "progressbar"
  )
})
names(fits) <- paste0("k", k_grid)

saveRDS(
  list(fits = fits, matrix_bundle = matrix_bundle, params = params),
  fit_rds
)
```

```{r inspect-topics, eval=FALSE}
library(fastTopics)

fit_bundle <- readRDS(fit_rds)
fits <- fit_bundle$fits
matrix_bundle <- fit_bundle$matrix_bundle

best_fit <- fits[[paste0("k", k_grid[1])]]

fastTopics::structure_plot(best_fit) +
  ggplot2::theme_classic() +
  ggplot2::labs(title = sprintf("%s fiber topics at %s sites", sample_name, paste(tf_names, collapse = ", ")))

topic_features <- as.data.table(best_fit$F, keep.rownames = "site_id")
topic_features_long <- melt(
  topic_features,
  id.vars = "site_id",
  variable.name = "topic",
  value.name = "feature_probability"
)

topic_features_long <- merge(
  topic_features_long,
  matrix_bundle$motif_meta,
  by = "site_id",
  all.x = TRUE
)

top_sites_per_topic <- topic_features_long[
  order(topic, -feature_probability),
  head(.SD, 25),
  by = topic
]

data.table::fwrite(
  topic_features_long,
  file.path(out_dir, sprintf("%s.%s.fastTopics_topic_site_loadings.tsv.gz", chrom, paste(tf_names, collapse = "_"))),
  sep = "\t"
)

top_sites_per_topic
```
