---
title: "Promoter-Level Fiber-seq Aggregation"
output:
  html_document:
    toc: true
    toc_float: true
---

This notebook builds a promoter-centered summary for the default sample and chromosome:

- `structure`: nucleosome features from `ft extract --nuc`
- `accessibility`: adenine modification calls from the phased BAM, summarized as m6A modified versus not-modified
- `methylation`: cytosine methylation calls from the phased BAM, summarized as 5mC modified versus not-modified

For scalability across all `chr1` promoters, the chromosome-wide accessibility and methylation summaries use `footprintR::readModBam(..., level = "summary")`. To keep the read-level adaptation from `/project/spott/cshan/code/fiberseqFunctions.R`, the notebook also generates one small example per-read binary m6A table for a single promoter window.


The threshold for defining whether a base is modified (`modProbThreshold`) is 0.9 because the score distribution is binary.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

required_pkgs <- c(
  "data.table",
  "dplyr",
  "tidyr",
  "tibble",
  "stringr",
  "GenomicRanges",
  "IRanges",
  "GenomeInfoDb",
  "SummarizedExperiment",
  "BiocParallel",
  "Rsamtools",
  "footprintR",
  "ggplot2",
  "patchwork",
  "scales",
  "signal"
)

missing_pkgs <- required_pkgs[
  !vapply(required_pkgs, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_pkgs) > 0) {
  stop(
    sprintf(
      "Missing required packages: %s",
      paste(missing_pkgs, collapse = ", ")
    ),
    call. = FALSE
  )
}

invisible(lapply(required_pkgs, library, character.only = TRUE))

default_sample <- "AL10_bc2178_19130"
default_chrom  <- "chr1"
window_bp <- 1e6L
bin_size <- 10L
mod_prob_threshold <- 0.9
example_tss_index <- 1L
results_root_dir <- "/project/spott/cshan/fiber-seq/results/promoter_summary/window_1Mb_TSS"
threshold_dir <- file.path(results_root_dir, "threshold_0.9")
ccre_output_dir <- file.path(results_root_dir, "ccre")

bam_path <- file.path(
  "/project/spott/1_Shared_projects/LCL_Fiber_seq/preprocess_final_merged_samples",
  sprintf("%s.5mC.6mA.aligned.phased.bam", default_sample)
)

tss_path <- "/project/spott/cshan/annotations/TSS.gencode.v49.bed"

sample_extract_dir <- file.path(
  "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results",
  default_sample,
  "extracted_results"
)

nuc_path <- file.path(
  sample_extract_dir,
  "nuc_by_chr",
  sprintf("%s.ft_extracted_nuc.%s.bed.gz", default_sample, default_chrom)
)

m6a_ft_path <- file.path(
  sample_extract_dir,
  "m6a_by_chr",
  sprintf("%s.ft_extracted_m6a.%s.bed.gz", default_sample, default_chrom)
)

cpg_ft_path <- file.path(
  sample_extract_dir,
  "cpg_by_chr",
  sprintf("%s.ft_extracted_cpg.%s.bed.gz", default_sample, default_chrom)
)

fire_peak_path <- file.path(
  "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results",
  default_sample,
  sprintf("%s-fire-v0.1-peaks.bed.gz", default_sample)
)

fire_element_path <- file.path(
  "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results",
  default_sample,
  "additional-outputs-v0.1",
  "fire-peaks",
  sprintf("%s-v0.1-fire-elements.bed.gz", default_sample)
)

output_dir <- file.path(threshold_dir, default_sample)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(ccre_output_dir, recursive = TRUE, showWarnings = FALSE)

bamfiles <- setNames(bam_path, default_sample)
bpparam_modbam <- BiocParallel::SerialParam()
```

## Helpers

```{r helpers}
assert_file_exists <- function(path, label = "file") {
  if (!file.exists(path)) {
    stop(sprintf("Missing %s: %s", label, path), call. = FALSE)
  }
  invisible(path)
}

write_gz_tsv <- function(x, path) {
  data.table::fwrite(x, file = path, sep = "\t")
  invisible(path)
}

read_bam_seqinfo <- function(bam_path) {
  targets <- Rsamtools::scanBamHeader(bam_path)[[1]]$targets
  GenomeInfoDb::Seqinfo(
    seqnames = names(targets),
    seqlengths = as.numeric(targets)
  )
}

read_gencode_tss <- function(path, chrom = NULL, window_bp = 2000L, seqinfo = NULL) {
  dt <- data.table::fread(
    path,
    header = FALSE,
    col.names = c("chromosome", "start0", "end0", "tss_id", "score", "tss_strand")
  )

  if (!is.null(chrom)) {
    dt <- dt[chromosome == chrom]
  }

  if (!nrow(dt)) {
    stop("No TSS rows remained after chromosome filtering.", call. = FALSE)
  }

  dt[, row_id := .I]
  dt[, tss_coordinate := start0 + 1L]
  dt[, gene_name_or_tss_id := tss_id]
  dt[, tss_uid := paste(chromosome, tss_coordinate, tss_strand, tss_id, row_id, sep = "|")]

  tss_gr <- GenomicRanges::GRanges(
    seqnames = dt$chromosome,
    ranges = IRanges::IRanges(dt$tss_coordinate, dt$tss_coordinate),
    strand = dt$tss_strand
  )

  if (!is.null(seqinfo)) {
    GenomeInfoDb::seqinfo(tss_gr) <- seqinfo[unique(dt$chromosome)]
  }

  mcols(tss_gr)$tss_uid <- dt$tss_uid
  mcols(tss_gr)$tss_id <- dt$tss_id
  mcols(tss_gr)$gene_name_or_tss_id <- dt$gene_name_or_tss_id

  promoter_windows <- GenomicRanges::promoters(
    tss_gr,
    upstream = window_bp,
    downstream = window_bp + 1L
  )

  promoter_windows <- GenomicRanges::trim(promoter_windows)

  promoter_meta <- data.table::data.table(
    tss_uid = dt$tss_uid,
    chromosome = dt$chromosome,
    tss_coordinate = dt$tss_coordinate,
    tss_strand = dt$tss_strand,
    gene_name_or_tss_id = dt$gene_name_or_tss_id,
    tss_id = dt$tss_id,
    promoter_start = start(promoter_windows),
    promoter_end = end(promoter_windows)
  )

  tss_table <- copy(promoter_meta)
  tss_table[, strand := tss_strand]
  tss_table[, tss_strand := NULL]
  data.table::setcolorder(
    tss_table,
    c(
      "chromosome",
      "tss_coordinate",
      "strand",
      "gene_name_or_tss_id",
      "tss_id",
      "tss_uid",
      "promoter_start",
      "promoter_end"
    )
  )

  list(
    tss_table = tss_table,
    tss_gr = tss_gr,
    promoter_windows = promoter_windows,
    promoter_meta = promoter_meta
  )
}

read_fiber_bed12 <- function(path) {
  data.table::fread(
    path,
    header = FALSE,
    col.names = c(
      "chrom",
      "chromStart",
      "chromEnd",
      "name",
      "score",
      "strand",
      "thickStart",
      "thickEnd",
      "itemRgb",
      "blockCount",
      "blockSizes",
      "blockStarts"
    )
  )
}

parse_bed12_blocks <- function(block_sizes, block_starts, drop_terminal_blocks = TRUE) {
  sizes <- as.integer(strsplit(block_sizes, ",", fixed = TRUE)[[1]])
  starts <- as.integer(strsplit(block_starts, ",", fixed = TRUE)[[1]])

  keep <- !is.na(sizes) & !is.na(starts)
  sizes <- sizes[keep]
  starts <- starts[keep]

  if (drop_terminal_blocks && length(sizes) >= 2L) {
    idx <- seq_along(sizes)
    idx <- idx[-c(1L, length(idx))]
    sizes <- sizes[idx]
    starts <- starts[idx]
  }

  list(sizes = sizes, starts = starts)
}

expand_ft_bed12_features <- function(bed12, feature_name = "feature", drop_terminal_blocks = TRUE) {
  feature_name_local <- feature_name
  
  # { ... } = run once per row/read
  # for each group, expand into multiple rows and return a new table using .()
  bed12[, {
    parsed <- parse_bed12_blocks(
      # block size = length of 1
      block_sizes = blockSizes, # a compressed string representation of multiple blocks for a single read
      block_starts = blockStarts,
      drop_terminal_blocks = drop_terminal_blocks
    )

    sizes <- parsed$sizes
    starts0 <- parsed$starts

    if (!length(sizes)) {
      return(NULL)
    }

    feature_start0 <- chromStart + starts0
    feature_end0 <- chromStart + starts0 + sizes
    feature_mid0 <- as.integer((feature_start0 + feature_end0) / 2)
 
    # .() is shorthand for list()
    .(
      feature_chromosome = chrom,
      read_id = name,
      read_strand = strand,
      feature_start0 = feature_start0,
      feature_end0 = feature_end0,
      feature_mid0 = feature_mid0,
      feature_pos = feature_mid0 + 1L,
      feature_size = sizes,
      feature_name = feature_name_local
    )
  }, by = .(chrom, chromStart, chromEnd, name, strand)]
}

# Converts feature table into GRanges
feature_blocks_to_granges <- function(feature_dt) {
  GenomicRanges::GRanges(
    seqnames = feature_dt$feature_chromosome,
    ranges = IRanges::IRanges(
      start = feature_dt$feature_start0 + 1L,
      end = feature_dt$feature_end0
    ),
    strand = "*"
  )
}


# Finds which features overlap TSS promoter windows and computes relative positions

## feature_dt: a data.table containing feature information, like chromosome, read ID, feature position, feature size
## feature_gr: the same features, but stored as a GRanges object for overlap operations
## promoter_windows: genomic promoter intervals around TSSs, also as GRanges
## promoter_meta: a table containing metadata for those promoter windows, such as TSS coordinate, strand, gene/TSS ID

align_feature_blocks_to_tss <- function(feature_dt, feature_gr, promoter_windows, promoter_meta) {
  if (!nrow(feature_dt) || !length(feature_gr) || !length(promoter_windows)) {
    # if any of the input is empty, return an empty data table instead of giving errors
    return(data.table::data.table())
  }

  # which feature block overlap which promoter windows?
  hits <- GenomicRanges::findOverlaps(
    # set query ranges to be the feature ranges --> intervals want to test for overlap
    query = feature_gr, 
    # subject ranges = promoter windows --> reference interval
    subject = promoter_windows,
    # overlap is only based on genommic coordinate
    ignore.strand = TRUE
  )

  if (!length(hits)) {
    return(data.table::data.table())
  }

  aligned <- cbind(
    # subset feature meta table to only include overlapping features
    feature_dt[S4Vectors::queryHits(hits)], 
    # subset promoter meta table to only include overlapping promoter windows
    promoter_meta[S4Vectors::subjectHits(hits)]
  )

  # Create a relative-position column based on strand
  ## add a new column to indicate relative position to TSS
  aligned[, rel_pos := ifelse(
    tss_strand == "+",
    # if + strand, relative position = feature - TSS coordinate --> positive number = donwstream
    feature_pos - tss_coordinate,
    tss_coordinate - feature_pos
  )]
  # return aligned data table after modifications
  aligned[]
}


# summarize nucleosome count, size and distance to TSS for each unique TSS/read
summarize_nucleosomes_by_tss <- function(nuc_aligned, window_bp = 2000L) {
  # takes output from  <align_feature_blocks_to_tss> to construct nuc_aligned
  if (!nrow(nuc_aligned)) {
    return(data.table::data.table(tss_uid = character()))
  }

  nuc_aligned[, {
    # .N = number of rows in the group
    n_nuc <- .N
    .(
      # n_nuc = nucleosome count per TSS
      structure_nucleosome_count = n_nuc,
      # average nucleosome size in this TSS group
      structure_mean_nucleosome_size = mean(feature_size),
      # Computes the median nucleosome size per TSS
      structure_median_nucleosome_size = as.double(median(feature_size)),
      # How many nucleosomes per kilobase are present around this TSS
      structure_nucleosome_density_per_kb = n_nuc / ((2L * window_bp + 1L) / 1000),
      # Counts how many nucleosomes have midpoint positions within 150 bp of the TSS
      structure_core_nucleosome_count = sum(abs(rel_pos) <= 150L),
      # Computes the average signed midpoint offset from the TSS.
      ## On average, are nucleosome midpoints shifted upstream or downstream of the TSS?
      structure_mean_midpoint_offset = mean(rel_pos),
      # How far away, on average, are nucleosome midpoints from the TSS?
      structure_mean_abs_midpoint_offset = mean(abs(rel_pos))
    )
  }, by = tss_uid]
}

## modbase_code == "a" ~ "m6A",
## modbase_code == "m" ~ "5mC"

# function that incorporates <footprintR::readModBam> to count how many bases were observed and how many were modified at each genomic position for this sample, then return the results as a clean table
read_modbam_summary_positions <- function(
  bamfiles,
  sample_name,
  promoter_windows,
  modbase_code,
  seqinfo, # A chromosome metadata object containing chromosome names and lengths
  bpparam,
  modProbThreshold = 0.9
) {
  se_summary <- footprintR::readModBam(
    bamfiles = bamfiles,
    regions = promoter_windows,
    modbase = setNames(modbase_code, sample_name), # Creates a named vector where the modification code is assigned to the sample name
    # aggregated counts at each genomic position
    ## count total valid bases per position/strand
    ## count modified bases per position/strand
    ## return assays Nvalid, Nmod, and FracM
    level = "summary",
    modProbThreshold = modProbThreshold,
    seqinfo = seqinfo,
    trim = TRUE,
    BPPARAM = bpparam,
    verbose = TRUE
  )
  
  # rows = genomic positions
  # columns = samples
  # assays = count summaries for those positions
  rr <- data.table::as.data.table(as.data.frame(SummarizedExperiment::rowRanges(se_summary)))
  rr <- rr[, .(
    call_chromosome = as.character(seqnames),
    position = start,
    call_strand = as.character(strand)
  )]
  
  # Pulls the "Nvalid" assay from the SummarizedExperiment
  ## Nvalid is the total number of valid/callable bases at each position for each sample
  total_calls <- as.integer(SummarizedExperiment::assay(se_summary, "Nvalid")[, sample_name])
  
  ## Nmod is the number of bases classified as modified at each position.
  modified_calls <- as.integer(SummarizedExperiment::assay(se_summary, "Nmod")[, sample_name])

  rr[, sample_name := sample_name]
  rr[, modbase_code := modbase_code]
  rr[, total_calls := total_calls]
  rr[, modified_calls := modified_calls]
  # Computes how many calls were not modified
  rr[, unmodified_calls := total_calls - modified_calls]
  rr[, frac_modified := data.table::fifelse(total_calls > 0, modified_calls / total_calls, NA_real_)]

  # only rows where there was at least one valid call
  rr[total_calls > 0]
}


# converts per-position modified vs unmodified counts into a long-format table with a status label and n_calls column
make_modified_status_table <- function(position_dt) {
  if (!nrow(position_dt)) {
    return(data.table::data.table())
  }

  modified_dt <- copy(position_dt)
  modified_dt[, `:=`(status = "modified", n_calls = modified_calls)]

  unmodified_dt <- copy(position_dt)
  unmodified_dt[, `:=`(status = "not_modified", n_calls = unmodified_calls)]

  data.table::rbindlist(
    list(modified_dt, unmodified_dt),
    use.names = TRUE,
    fill = TRUE
  )[n_calls > 0]
}

align_positions_to_tss <- function(position_dt, promoter_windows, promoter_meta) {
  if (!nrow(position_dt) || !length(promoter_windows)) {
    return(data.table::data.table())
  }

  pos_gr <- GenomicRanges::GRanges(
    seqnames = position_dt$call_chromosome,
    ranges = IRanges::IRanges(position_dt$position, position_dt$position),
    strand = "*"
  )

  hits <- GenomicRanges::findOverlaps(
    query = pos_gr,
    subject = promoter_windows,
    ignore.strand = TRUE
  )

  if (!length(hits)) {
    return(data.table::data.table())
  }

  aligned <- cbind(
    position_dt[S4Vectors::queryHits(hits)],
    promoter_meta[S4Vectors::subjectHits(hits)]
  )

  aligned[, rel_pos := ifelse(
    tss_strand == "+",
    position - tss_coordinate,
    tss_coordinate - position
  )]

  aligned[]
}

summarize_modality_by_tss <- function(aligned_dt, prefix = c("accessibility", "methylation")) {
  prefix <- match.arg(prefix)

  if (!nrow(aligned_dt)) {
    return(data.table::data.table(tss_uid = character()))
  }

  out <- aligned_dt[, {
    total <- sum(total_calls)
    modified <- sum(modified_calls)
    unmodified <- sum(unmodified_calls)
    .(
      modified_calls = modified,
      unmodified_calls = unmodified,
      total_calls = total,
      positions_with_coverage = sum(total_calls > 0),
      fraction_modified = if (total > 0) modified / total else NA_real_
    )
  }, by = tss_uid]

  data.table::setnames(
    out,
    old = c("modified_calls", "unmodified_calls", "total_calls", "positions_with_coverage", "fraction_modified"),
    new = c(
      sprintf("%s_modified_calls", prefix),
      sprintf("%s_unmodified_calls", prefix),
      sprintf("%s_total_calls", prefix),
      sprintf("%s_positions_with_coverage", prefix),
      sprintf("%s_fraction_modified", prefix)
    )
  )

  out[]
}

make_nucleosome_meta <- function(nuc_aligned, num_tss, window_bp = 2000L, bin_size = 10L) {
  bins <- data.table::data.table(bin = seq(-window_bp, window_bp, by = bin_size))

  if (!nrow(nuc_aligned)) {
    bins[, nucleosome_midpoints_per_tss := 0]
    return(bins[])
  }

  tmp <- copy(nuc_aligned)
  tmp[, bin := round(rel_pos / bin_size) * bin_size]

  meta <- tmp[, .N, by = bin]
  meta <- merge(bins, meta, by = "bin", all.x = TRUE, sort = TRUE)
  meta[is.na(N), N := 0]
  meta[, nucleosome_midpoints_per_tss := N / num_tss]
  meta[]
}

# takes modification calls already aligned relative to TSSs, groups them into position bins, summarizes modified and total calls per bin, computes fraction modified and modified calls per TSS

make_modification_meta <- function(aligned_dt, num_tss, label, window_bp = 2000L, bin_size = 10L) {
  bins <- data.table::data.table(bin = seq(-window_bp, window_bp, by = bin_size))
  label_value <- label

  if (!nrow(aligned_dt)) {
    bins[, `:=`(
      modified_calls = 0L,
      total_calls = 0L,
      fraction_modified = NA_real_,
      modified_calls_per_tss = 0,
      label = label_value
    )]
    return(bins[])
  }

  tmp <- copy(aligned_dt)
  # converts each relative position rel_pos into the nearest bin value
  tmp[, bin := round(rel_pos / bin_size) * bin_size]

  meta <- tmp[, .(
    modified_calls = sum(modified_calls),
    total_calls = sum(total_calls)
  ), by = bin]

  meta <- merge(bins, meta, by = "bin", all.x = TRUE, sort = TRUE)
  meta[is.na(modified_calls), modified_calls := 0L]
  meta[is.na(total_calls), total_calls := 0L]
  meta[, fraction_modified := data.table::fifelse(total_calls > 0, modified_calls / total_calls, NA_real_)]
  meta[, modified_calls_per_tss := modified_calls / num_tss]
  meta[, label := label_value]
  meta[]
}

plot_structure_meta <- function(
  meta_dt,
  sample_name,
  chrom,
  window_bp        = 5000L,   # x-axis range (+/- bp from TSS)
  x_break_interval = 500,     # spacing of x-axis tick marks (bp)
  loess_span       = 0.06,    # LOESS smoothing bandwidth (smaller = tighter fit)
  ylim             = NULL,    # c(lo, hi) to fix y-axis, NULL for auto
  line_color       = "#4D4D4D",
  smooth_color     = "#1f77b4",
  linewidth        = 0.6,     # raw data line width
  smooth_linewidth = 0.9,     # LOESS line width
  base_size        = 12       # ggplot theme base font size
) {
  p <- ggplot2::ggplot(meta_dt, ggplot2::aes(x = bin, y = nucleosome_midpoints_per_tss)) +
    ggplot2::geom_line(color = line_color, linewidth = linewidth) +
    ggplot2::geom_smooth(
      method = "loess",
      span   = loess_span,
      se     = FALSE,
      color  = smooth_color,
      linewidth = smooth_linewidth
    ) +
    ggplot2::scale_x_continuous(
      breaks = seq(-window_bp, window_bp, x_break_interval),
      labels = function(x) paste0(x / 1000, " kb")
    ) +
    ggplot2::geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.4) +
    ggplot2::labs(
      x     = "Distance to TSS",
      y     = "Nucleosome midpoints per TSS",
      title = sprintf("Promoter structure: %s %s", sample_name, chrom)
    ) +
    ggplot2::theme_bw(base_size = base_size)

  if (!is.null(ylim)) {
    p <- p + ggplot2::coord_cartesian(ylim = ylim)
  }
  p
}

plot_modification_meta <- function(
  meta_dt,
  title,
  color,
  window_bp        = 5000L,   # x-axis range (+/- bp from TSS)
  x_break_interval = 500,     # spacing of x-axis tick marks (bp)
  loess_span       = 0.06,    # LOESS smoothing bandwidth (smaller = tighter fit)
  ylim             = NULL,    # c(lo, hi) to fix y-axis, NULL for auto
  linewidth        = 0.6,     # raw data line width
  smooth_linewidth = 0.9,     # LOESS line width
  y_label          = "Modified fraction",
  base_size        = 12       # ggplot theme base font size
) {
  p <- ggplot2::ggplot(meta_dt, ggplot2::aes(x = bin, y = fraction_modified)) +
    ggplot2::geom_line(color = color, linewidth = linewidth, na.rm = TRUE) +
    ggplot2::geom_smooth(
      method = "loess",
      span   = loess_span,
      se     = FALSE,
      color  = color,
      linewidth = smooth_linewidth,
      na.rm  = TRUE
    ) +
    ggplot2::scale_x_continuous(
      breaks = seq(-window_bp, window_bp, x_break_interval),
      labels = function(x) paste0(x / 1000, " kb")
    ) +
    ggplot2::geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 0.4) +
    ggplot2::labs(
      x     = "Distance to TSS",
      y     = y_label,
      title = title
    ) +
    ggplot2::theme_bw(base_size = base_size)

  if (!is.null(ylim)) {
    p <- p + ggplot2::coord_cartesian(ylim = ylim)
  }
  p
}

# Adapted from /project/spott/cshan/code/fiberseqFunctions.R

## takes a SummarizedExperiment containing read-level modification probabilities, converts those probabilities into 0/1 modified calls using a threshold, reshapes the data into a long table with one row per genomic position per read, and annotates each row with chromosome, position, strand, sample, and modification label
getBinarizedModificationsLongTable <- function(footprintRObject, sampleName, modProbThreshold = 0.9) {
  if (!inherits(footprintRObject, "SummarizedExperiment")) {
    stop("footprintRObject must be a SummarizedExperiment.", call. = FALSE)
  }

  # default is at per read level
  ## Extracts the assay named "mod_prob" from the SummarizedExperiment
  mod_prob_assay <- SummarizedExperiment::assay(footprintRObject, "mod_prob")[[sampleName]]
  mod_prob_matrix <- as(mod_prob_assay, "matrix")

  # 1 = probability greater than threshold = modified 
  mod_prob_matrix[mod_prob_matrix > modProbThreshold] <- 1
  mod_prob_matrix[mod_prob_matrix <= modProbThreshold] <- 0

  rownames(mod_prob_matrix) <- rownames(footprintRObject)
  mod_wide <- tibble::as_tibble(mod_prob_matrix, rownames = "genomic_location")

  # Reshapes from wide to long format
  mod_long <- tidyr::pivot_longer(
    mod_wide,
    cols = -genomic_location,
    names_to = "read_id",
    values_to = "mod_status"
  )

  # Parse genomic location and annotate call state
  mod_long <- dplyr::mutate(
    mod_long,
    call_chromosome = stringr::str_extract(genomic_location, "^[^:]+"),
    position = as.integer(stringr::str_extract(genomic_location, "(?<=:)[0-9]+")),
    call_strand = stringr::str_extract(genomic_location, "[+-]$"),
    sample_name = sampleName,
    call_state = ifelse(mod_status == 1, "modified", "not_modified")
  )

  modbase_code <- SummarizedExperiment::colData(footprintRObject)$modbase[[sampleName]]

  mod_long <- dplyr::mutate(
    mod_long,
    modbase_code = modbase_code,
    modbase_label = dplyr::case_when(
      modbase_code == "a" ~ "m6A",
      modbase_code == "m" ~ "5mC",
      TRUE ~ modbase_code
    )
  )

  dplyr::select(
    mod_long,
    sample_name,
    modbase_code,
    modbase_label,
    read_id,
    call_chromosome,
    position,
    call_strand,
    mod_status,
    call_state
  )
}

# Make a vector with 20 positive weights, then 120 negative weights, then 20 positive weights
build_nucleosome_weight_vector <- function() {
  c(
    rep(0.375, 20),
    rep(-0.125, 120),
    rep(0.375, 20)
  )
}

make_chapter5_example_plot <- function(
  bamfiles,
  sample_name,
  seqinfo,
  promoter_windows,
  promoter_meta,
  mod_prob_threshold,
  example_tss_index,
  output_dir,
  chrom
) {
  if (length(promoter_windows) < example_tss_index) {
    return(NULL)
  }

  example_region <- promoter_windows[example_tss_index]
  example_tss_uid <- promoter_meta$tss_uid[example_tss_index]

  se_example <- footprintR::readModBam(
    bamfiles = bamfiles,
    regions = example_region,
    modbase = setNames("a", sample_name),
    level = "quickread",
    seqinfo = seqinfo,
    trim = TRUE,
    BPPARAM = BiocParallel::SerialParam(),
    verbose = TRUE
  )

  se_example <- footprintR::flattenReadLevelAssay(se_example)
  wgt <- build_nucleosome_weight_vector()
  se_example <- footprintR::addFootprints(
    se = se_example,
    wgt = wgt,
    thresh = 0.02,
    name = "nucl"
  )

  example_binary_calls <- getBinarizedModificationsLongTable(
    footprintRObject = se_example,
    sampleName = sample_name,
    modProbThreshold = mod_prob_threshold
  )

  example_binary_calls <- data.table::as.data.table(example_binary_calls)
  example_binary_calls[, example_tss_uid := example_tss_uid]
  write_gz_tsv(
    example_binary_calls,
    file.path(
      output_dir,
      sprintf("%s_%s_exampleTSS_read_level_m6A_binary_calls.tsv.gz", sample_name, chrom)
    )
  )

  p <- footprintR::plotRegion(
    se_example,
    region = example_region,
    minCoveredFraction = 0,
    tracks = list(
      list(
        trackData = "mod_prob",
        trackType = "Heatmap",
        interpolate = TRUE,
        footprintColumns = "nucl",
        arglistFootprints = list(
          nucl = list(color = scales::alpha("firebrick", 0.5))
        ),
        orderReads = "squish"
      ),
      list(
        trackData = "FracMod",
        trackType = "Smooth",
        smoothMethod = "rollingMean",
        windowSize = 41
      )
    )
  ) + patchwork::plot_layout(heights = c(2, 0.6))

  ggplot2::ggsave(
    filename = file.path(
      output_dir,
      sprintf("%s_%s_exampleTSS_footprintR_nucleosome_plot.pdf", sample_name, chrom)
    ),
    plot = p,
    width = 10,
    height = 6
  )

  list(
    plot = p,
    binary_calls = example_binary_calls,
    example_tss_uid = example_tss_uid
  )
}
```

## Validate inputs

Check that `run_modbam_promoter_summary.R` has already been run and output files exist.

```{r validate-inputs}
expected_files <- c(
  file.path(output_dir, sprintf("%s_%s_tss_table.tsv.gz",                                          default_sample, default_chrom)),
  file.path(output_dir, sprintf("%s_%s_promoter_structure.tsv.gz",                                 default_sample, default_chrom)),
  file.path(output_dir, sprintf("%s_%s_nucleosome_meta.tsv.gz",                                    default_sample, default_chrom)),
  file.path(output_dir, sprintf("%s_%s_promoter_accessibility.tsv.gz",                             default_sample, default_chrom)),
  file.path(output_dir, sprintf("%s_%s_m6A_meta.tsv.gz",                                           default_sample, default_chrom)),
  file.path(output_dir, sprintf("%s_%s_promoter_methylation.tsv.gz",                               default_sample, default_chrom)),
  file.path(output_dir, sprintf("%s_%s_5mC_meta.tsv.gz",                                           default_sample, default_chrom)),
  file.path(output_dir, sprintf("%s_%s_promoter_structure_accessibility_methylation.tsv.gz",       default_sample, default_chrom))
)

missing <- expected_files[!file.exists(expected_files)]
if (length(missing)) {
  stop(
    "Missing pre-computed files. Run run_modbam_promoter_summary.R first:\n",
    paste(missing, collapse = "\n"),
    call. = FALSE
  )
}
message("All expected output files found.")
```

## TSS table and promoter windows

```{r tss-table}
tss_table <- data.table::fread(
  file.path(output_dir, sprintf("%s_%s_tss_table.tsv.gz", default_sample, default_chrom))
)

tss_table[1:10]
```

## Promoter structure from `ft extract --nuc`

```{r promoter-structure}
nuc_summary <- data.table::fread(
  file.path(output_dir, sprintf("%s_%s_promoter_structure.tsv.gz", default_sample, default_chrom))
)

nuc_meta <- data.table::fread(
  file.path(output_dir, sprintf("%s_%s_nucleosome_meta.tsv.gz", default_sample, default_chrom))
)

nuc_summary[1:10]
```

## Promoter accessibility from BAM m6A calls

Results pre-computed by `run_modbam_promoter_summary.R` (submit via
`code/run_modbam_promoter_summary.sh`). Read output files below.

```{r promoter-accessibility}
m6a_summary <- data.table::fread(
  file.path(output_dir, sprintf("%s_%s_promoter_accessibility.tsv.gz", default_sample, default_chrom))
)

m6a_meta <- data.table::fread(
  file.path(output_dir, sprintf("%s_%s_m6A_meta.tsv.gz", default_sample, default_chrom))
)

m6a_summary[1:10]
```

## Promoter methylation from BAM 5mC calls

```{r promoter-methylation}
m5c_summary <- data.table::fread(
  file.path(output_dir, sprintf("%s_%s_promoter_methylation.tsv.gz", default_sample, default_chrom))
)

m5c_meta <- data.table::fread(
  file.path(output_dir, sprintf("%s_%s_5mC_meta.tsv.gz", default_sample, default_chrom))
)

m5c_summary[1:10]
```

## Join structure, accessibility, and methylation per promoter

```{r promoter-join}
# Pre-computed and saved by run_modbam_promoter_summary.R
promoter_multiomic_summary <- data.table::fread(
  file.path(
    output_dir,
    sprintf("%s_%s_promoter_structure_accessibility_methylation.tsv.gz", default_sample, default_chrom)
  )
)

promoter_multiomic_summary[1:10]
```

## Promoter-centered metaprofiles

```{r promoter-metaprofiles, fig.width=16, fig.height=4.5}
# load nuc_meta, m6a_meta, and m5c_meta from storage
# /project/spott/cshan/fiber-seq/results/promoter_summary/AL10_bc2178_19130
source("/project/spott/cshan/fiber-seq/code/footprintR_modbam_code/modbam_footprintR_functions.R")

output_dir <- "/project/spott/cshan/fiber-seq/results/promoter_summary/5kb_TSS/threshold_0.9/AL10_bc2178_19130"
default_sample <- "AL10_bc2178_19130"
default_chrom <- "chr1"

nuc_meta <- data.table::fread(
  file.path(output_dir, sprintf("%s_%s_nucleosome_meta.tsv.gz", default_sample, default_chrom))
)
m6a_meta <- data.table::fread(
  file.path(output_dir, sprintf("%s_%s_m6A_meta.tsv.gz", default_sample, default_chrom))
)


m5c_meta <- data.table::fread(
  file.path(output_dir, sprintf("%s_%s_5mC_meta.tsv.gz", default_sample, default_chrom))
)


structure_plot <- plot_structure_meta(
  meta_dt         = nuc_meta,
  sample_name     = default_sample,
  chrom           = default_chrom,
  window_bp        = 5000L,       # x-axis range (+/- bp)
  x_break_interval = 500,         # tick spacing (bp)
  loess_span       = 0.06,        # smoothing bandwidth
  ylim             = NULL,        # e.g. c(0, 0.5) to fix y-axis
  line_color       = "#4D4D4D",
  smooth_color     = "#1f77b4",
  linewidth        = 0.6,
  smooth_linewidth = 0.9,
  base_size        = 12
)

accessibility_plot <- plot_modification_meta(
  meta_dt          = m6a_meta,
  title            = sprintf("Promoter accessibility: %s %s", default_sample, default_chrom),
  color            = "#7B3294",
  window_bp        = 5000L,       # x-axis range (+/- bp)
  x_break_interval = 500,         # tick spacing (bp)
  loess_span       = 0.06,        # smoothing bandwidth
  ylim             = NULL,        # e.g. c(0, 0.3) to fix y-axis
  linewidth        = 0.6,
  smooth_linewidth = 0.9,
  y_label          = "Modified fraction",
  base_size        = 12
)

methylation_plot <- plot_modification_meta(
  meta_dt          = m5c_meta,
  title            = sprintf("Promoter methylation: %s %s", default_sample, default_chrom),
  color            = "#8C510A",
  window_bp        = 5000L,       # x-axis range (+/- bp)
  x_break_interval = 500,         # tick spacing (bp)
  loess_span       = 0.06,        # smoothing bandwidth
  ylim             = NULL,        # e.g. c(0, 0.3) to fix y-axis
  linewidth        = 0.6,
  smooth_linewidth = 0.9,
  y_label          = "Modified fraction",
  base_size        = 12
)

# Overlay the three plots using patchwork



# Overlay the plots using patchwork
# structure_plot is included as-is, which allows it to autoscale
combined_plot <- structure_plot / accessibility_plot / methylation_plot + 
  patchwork::plot_layout(heights = c(1, 1, 1))

# Save the plot
ggplot2::ggsave(
  filename = file.path(
    output_dir,
    sprintf("%s_%s_promoter_multiomic_metaprofiles_5kb_threshold_0.9.pdf", default_sample, default_chrom)
  ),
  plot = combined_plot,
  width = 10,
  height = 16
)

# Display the plot
combined_plot
```

## One example `footprintR` nucleosome call from the BAM

This small example follows the chapter-5 `footprintR` nucleosome workflow on a single promoter window. It keeps the notebook tied to the `nucleosomes.html` analysis pattern while the chromosome-wide summary stays scalable.

```{r chapter5-example, fig.width=10, fig.height=6}
# use example promoter region from region_str = "chr3:186782192-186785192"
test_window =   GenomicRanges::GRanges(
  seqnames = default_chrom,
  ranges = IRanges::IRanges(186782192, 186785192),
  strand = "+"
)

# extract promoter_meta to match test_window
test_tss_uid <- promoter_meta$tss_uid[1] 
test_promoter_meta <- promoter_meta[tss_uid == test_tss_uid]

# get example_tss_index by matching test_tss_uid to promoter_meta$tss_uid
example_tss_index <- which(promoter_meta$tss_uid == test_tss_uid)

example_footprint <- make_chapter5_example_plot(
  bamfiles = bamfiles,
  sample_name = default_sample,
  seqinfo = bam_seqinfo,
  promoter_windows = test_window,
  promoter_meta = test_promoter_meta,
  mod_prob_threshold = mod_prob_threshold,
  example_tss_index = example_tss_index,
  output_dir = output_dir,
  chrom = default_chrom
)

if (!is.null(example_footprint)) {
  example_footprint$plot
}
```

## Output files

```{r output-files}
list.files(output_dir, full.names = TRUE)
```

# Examine % modified

```{r}
library(data.table)
library(ggplot2)
library(tidyr)

# 1. Define the file path (using the variables from your environment)
saved_file <- file.path(
  output_dir,
  sprintf("%s_%s_promoter_structure_accessibility_methylation.tsv.gz", default_sample, default_chrom)
)

# 2. Load the data
summary_data <- data.table::fread(saved_file)

# 3. Prepare data for QC Bar Plot (Calculating global means)
qc_summary <- data.frame(
  Modality = c("Accessibility (m6A)", "Methylation (5mC)"),
  Mean_Fraction = c(
    mean(summary_data$accessibility_fraction_modified, na.rm = TRUE),
    mean(summary_data$methylation_fraction_modified, na.rm = TRUE)
  )
)

# 4. Create the QC Bar Plot
qc_plot <- ggplot(qc_summary, aes(x = Modality, y = Mean_Fraction, fill = Modality)) +
  geom_bar(stat = "identity", color = "black", width = 0.6) +
  scale_fill_manual(values = c("Accessibility (m6A)" = "#7B3294", "Methylation (5mC)" = "#008837")) +
  scale_y_continuous(limits = c(0, 1), expand = c(0, 0)) +
  labs(
    title = paste("QC: Average Promoter Signal -", default_sample),
    subtitle = paste("Region:", default_chrom),
    y = "Mean Fraction Modified",
    x = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Display the plot
print(qc_plot)

# Histogram to see the distribution across all promoters
dist_plot <- summary_data %>%
  pivot_longer(cols = c(accessibility_fraction_modified, methylation_fraction_modified),
               names_to = "Modality", values_to = "Fraction") %>%
  ggplot(aes(x = Fraction, fill = Modality)) +
  geom_histogram(bins = 50, alpha = 0.7, position = "identity") +
  theme_minimal() +
  labs(title = "Distribution of Modifications across Promoters")

print(dist_plot)

plot_dir <- "/project/spott/cshan/fiber-seq/results/promoter_summary/threshold_0.9/AL10_bc2178_19130/QC_plots/"

combined_QC_plot <- qc_plot / dist_plot + patchwork::plot_layout(heights = c(1, 1))
combined_QC_plot <- ggarrange(qc_plot, dist_plot, ncol = 1, nrow = 2)
ggsave(file.path(paste0(plot_dir, "accessibility_methylation_dist.png")), combined_QC_plot)
```

Correlation Plot (Accessibility vs. Methylation)
Check if individual promoters follow the expected rule: High Accessibility = Low Methylation
```{r}

pdf(file.path(paste0(plot_dir, "accessibility_methylation_correlation.pdf")))
ggplot(summary_data, aes(x = accessibility_fraction_modified, y = methylation_fraction_modified)) +
  geom_point(alpha = 0.2, color = "darkblue") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Promoter-wise: Accessibility vs. Methylation",
       x = "m6A Fraction (Accessibility)", y = "5mC Fraction (Methylation)")
dev.off()


```



Does having more nucleosomes at the promoter actually decrease the measured accessibility?
```{r}

pdf(file.path(paste0(plot_dir, "accessibility_nuc.pdf")), width = 12)
ggplot(summary_data, aes(x = factor(structure_core_nucleosome_count), y = accessibility_fraction_modified)) +
  geom_boxplot(fill = "#7B3294", alpha = 0.6) +
  labs(title = "Accessibility by Nucleosome Count",
       x = "Number of Nucleosomes in Promoter Core", y = "m6A Fraction")
dev.off()

```

Coverage QC (Total Calls and modified Distribution)

```{r}

ggplot(summary_data, aes(x = accessibility_total_calls)) +
  geom_histogram(bins = 100, fill = "steelblue") +
  labs(title = "Distribution of Read Coverage per Promoter",
       x = "Total Valid Calls", y = "Count of Promoters") 
```



# Examine cCRE to examine their m6A and 5mc distribution
##############################################################

  - PLS = promoter like
  - pELS = proximial enhancer like
  - dELS = distal enhancer like
  - chromatin accessible with TF = CA-TF
  - chromatin accessible with CTCF = CA-CTC
  - chromatin accessible with H3K4me3 = CA-H3K4me3
  - all candidate enhancers = pELS + dELS
  - chromatin accessible only = CA
  - TF only = TF
  

  https://screen.wenglab.org/downloads

  donwload cCRE is at /project/spott/cshan/annotations/GRCh38-cCREs.bed

```{bash, eval = FALSE}
# donwload  cCREs

wget https://downloads.wenglab.org/Registry-V4/GRCh38-cCREs.bed

```



## Random-region and cCRE sanity check

The goal of this section is to check the m6A and 5mC signal distribution in random genomic regions and cCREs, to confirm that the promoter patterns we see are specific to promoters and not a genome-wide artifact

```{r random-region-cCRE, fig.width=12, fig.height=10}
# ── Variables ────────────────────────────────────────────────────────────────
default_sample     <- "AL10_bc2178_19130"
default_chrom      <- "chr1"
window_bp          <- 2000L
mod_prob_threshold <- 0.9

results_root_dir <- "/project/spott/cshan/fiber-seq/results/promoter_summary"
ccre_output_dir  <- file.path(results_root_dir, "ccre")
dir.create(ccre_output_dir, recursive = TRUE, showWarnings = FALSE)

bam_path <- file.path(
  "/project/spott/1_Shared_projects/LCL_Fiber_seq/preprocess_final_merged_samples",
  sprintf("%s.5mC.6mA.aligned.phased.bam", default_sample)
)
tss_path    <- "/project/spott/cshan/annotations/TSS.gencode.v49.bed"
cpg_ft_path <- file.path(
  "/project/spott/1_Shared_projects/LCL_Fiber_seq/FIRE/results",
  default_sample, "extracted_results", "cpg_by_chr",
  sprintf("%s.ft_extracted_cpg.%s.bed.gz", default_sample, default_chrom)
)
ccre_path <- "/project/spott/cshan/annotations/GRCh38-cCREs.bed"

# ── Utility functions (mirrors Helpers chunk) ─────────────────────────────────
assert_file_exists <- function(path, label = "file") {
  if (!file.exists(path)) stop(sprintf("Missing %s: %s", label, path), call. = FALSE)
  invisible(path)
}

write_gz_tsv <- function(x, path) {
  data.table::fwrite(x, file = path, sep = "\t")
  invisible(path)
}

read_bam_seqinfo <- function(bam_path) {
  targets <- Rsamtools::scanBamHeader(bam_path)[[1]]$targets
  GenomeInfoDb::Seqinfo(seqnames = names(targets), seqlengths = as.numeric(targets))
}

read_gencode_tss <- function(path, chrom = NULL, window_bp = 2000L, seqinfo = NULL) {
  dt <- data.table::fread(path, header = FALSE,
    col.names = c("chromosome", "start0", "end0", "tss_id", "score", "tss_strand"))
  if (!is.null(chrom)) dt <- dt[chromosome == chrom]
  if (!nrow(dt)) stop("No TSS rows remained after chromosome filtering.", call. = FALSE)
  dt[, tss_coordinate     := start0 + 1L]
  dt[, tss_uid            := paste(chromosome, tss_coordinate, tss_strand, tss_id, .I, sep = "|")]
  tss_gr <- GenomicRanges::GRanges(
    seqnames = dt$chromosome,
    ranges   = IRanges::IRanges(dt$tss_coordinate, dt$tss_coordinate),
    strand   = dt$tss_strand
  )
  if (!is.null(seqinfo)) GenomeInfoDb::seqinfo(tss_gr) <- seqinfo[unique(dt$chromosome)]
  list(
    tss_gr           = tss_gr,
    promoter_windows = GenomicRanges::trim(
      GenomicRanges::promoters(tss_gr, upstream = window_bp, downstream = window_bp + 1L)
    )
  )
}

read_fiber_bed12 <- function(path) {
  data.table::fread(path, header = FALSE,
    col.names = c("chrom","chromStart","chromEnd","name","score","strand",
                  "thickStart","thickEnd","itemRgb","blockCount","blockSizes","blockStarts"))
}

parse_bed12_blocks <- function(block_sizes, block_starts, drop_terminal_blocks = TRUE) {
  sizes  <- as.integer(strsplit(block_sizes,  ",", fixed = TRUE)[[1]])
  starts <- as.integer(strsplit(block_starts, ",", fixed = TRUE)[[1]])
  keep   <- !is.na(sizes) & !is.na(starts)
  sizes  <- sizes[keep];  starts <- starts[keep]
  if (drop_terminal_blocks && length(sizes) >= 2L) {
    idx <- seq_along(sizes)[-c(1L, length(sizes))]
    sizes <- sizes[idx];  starts <- starts[idx]
  }
  list(sizes = sizes, starts = starts)
}

expand_ft_bed12_features <- function(bed12, drop_terminal_blocks = TRUE) {
  bed12[, {
    parsed        <- parse_bed12_blocks(blockSizes, blockStarts, drop_terminal_blocks)
    if (!length(parsed$sizes)) return(NULL)
    feature_start0 <- chromStart + parsed$starts
    feature_end0   <- chromStart + parsed$starts + parsed$sizes
    feature_mid0   <- as.integer((feature_start0 + feature_end0) / 2)
    .(feature_chromosome = chrom,
      feature_start0 = feature_start0, feature_end0 = feature_end0,
      feature_mid0   = feature_mid0,   feature_pos   = feature_mid0 + 1L,
      feature_size   = parsed$sizes)
  }, by = .(chrom, chromStart, chromEnd, name, strand)]
}

feature_blocks_to_granges <- function(feature_dt) {
  GenomicRanges::GRanges(
    seqnames = feature_dt$feature_chromosome,
    ranges   = IRanges::IRanges(feature_dt$feature_start0 + 1L, feature_dt$feature_end0),
    strand   = "*"
  )
}

read_modbam_summary_positions <- function(bamfiles, sample_name, promoter_windows,
                                          modbase_code, seqinfo, bpparam,
                                          modProbThreshold = 0.9) {
  se_summary <- footprintR::readModBam(
    bamfiles         = bamfiles,
    regions          = promoter_windows,
    modbase          = setNames(modbase_code, sample_name),
    level            = "summary",
    modProbThreshold = modProbThreshold,
    seqinfo          = seqinfo,
    trim             = TRUE,
    BPPARAM          = bpparam,
    verbose          = FALSE
  )
  rr <- data.table::as.data.table(as.data.frame(SummarizedExperiment::rowRanges(se_summary)))
  rr <- rr[, .(call_chromosome = as.character(seqnames),
               position        = start,
               call_strand     = as.character(strand))]
  total_calls    <- as.integer(SummarizedExperiment::assay(se_summary, "Nvalid")[, sample_name])
  modified_calls <- as.integer(SummarizedExperiment::assay(se_summary, "Nmod")[,  sample_name])
  rr[, `:=`(sample_name      = sample_name,
             modbase_code     = modbase_code,
             total_calls      = total_calls,
             modified_calls   = modified_calls,
             unmodified_calls = total_calls - modified_calls,
             frac_modified    = data.table::fifelse(total_calls > 0,
                                  modified_calls / total_calls, NA_real_))]
  rr[total_calls > 0]
}

assert_file_exists(bam_path,     "BAM")
assert_file_exists(cpg_ft_path,  "CpG ft extract BED")
assert_file_exists(ccre_path,    "cCRE BED")

bamfiles      <- setNames(bam_path, default_sample)
bpparam_modbam <- BiocParallel::SerialParam()

# --- New helper  ---

region_granges <- function(region_row) {
  GenomicRanges::GRanges(
    seqnames = region_row$chromosome,
    ranges   = IRanges::IRanges(region_row$start0 + 1L, region_row$end0),
    strand   = "*"
  )
}

summarize_region_modbam <- function(region_row, bamfiles, sample_name, seqinfo, bpparam, modbase_code) {
  position_dt <- read_modbam_summary_positions(
    bamfiles         = bamfiles,
    sample_name      = sample_name,
    promoter_windows = region_granges(region_row),
    modbase_code     = modbase_code,
    seqinfo          = seqinfo,
    bpparam          = bpparam,
    modProbThreshold = mod_prob_threshold
  )
  total_calls    <- sum(position_dt$total_calls)
  modified_calls <- sum(position_dt$modified_calls)
  data.table::data.table(
    region_source           = region_row$region_source,
    region_class            = region_row$region_class,
    region_label            = region_row$region_label,
    chromosome              = region_row$chromosome,
    start                   = region_row$start0 + 1L,
    end                     = region_row$end0,
    region_width_bp         = region_row$end0 - region_row$start0,
    modbase_code            = modbase_code,
    modified_calls          = modified_calls,
    unmodified_calls        = sum(position_dt$unmodified_calls),
    total_calls             = total_calls,
    positions_with_coverage = sum(position_dt$total_calls > 0),
    fraction_modified       = if (total_calls > 0) modified_calls / total_calls else NA_real_
  )
}

summarize_region_cpg <- function(region_row, cpg_bed12_all) {
  region_gr     <- region_granges(region_row)
  cpg_candidate <- cpg_bed12_all[
    chrom      == region_row$chromosome &
      chromEnd   >= region_row$start0 + 1L &
      chromStart <= region_row$end0
  ]
  empty_row <- data.table::data.table(
    region_source             = region_row$region_source,
    region_class              = region_row$region_class,
    region_label              = region_row$region_label,
    chromosome                = region_row$chromosome,
    start                     = region_row$start0 + 1L,
    end                       = region_row$end0,
    region_width_bp           = region_row$end0 - region_row$start0,
    cpg_feature_blocks        = 0L,
    cpg_feature_blocks_per_kb = 0,
    cpg_covered_bases         = 0L,
    cpg_covered_fraction      = 0
  )
  if (!nrow(cpg_candidate)) return(empty_row)
  cpg_features <- expand_ft_bed12_features(cpg_candidate, drop_terminal_blocks = FALSE)
  if (!nrow(cpg_features)) return(empty_row)
  cpg_gr   <- feature_blocks_to_granges(cpg_features)
  overlaps <- GenomicRanges::findOverlaps(cpg_gr, region_gr, ignore.strand = TRUE)
  if (!length(overlaps)) return(empty_row)
  hit_idx       <- unique(S4Vectors::queryHits(overlaps))
  covered_bases <- sum(width(GenomicRanges::pintersect(cpg_gr[hit_idx], region_gr)))
  cpg_blocks    <- length(hit_idx)
  region_width  <- region_row$end0 - region_row$start0
  data.table::data.table(
    region_source             = region_row$region_source,
    region_class              = region_row$region_class,
    region_label              = region_row$region_label,
    chromosome                = region_row$chromosome,
    start                     = region_row$start0 + 1L,
    end                       = region_row$end0,
    region_width_bp           = region_width,
    cpg_feature_blocks        = cpg_blocks,
    cpg_feature_blocks_per_kb = cpg_blocks / (region_width / 1000),
    cpg_covered_bases         = covered_bases,
    cpg_covered_fraction      = covered_bases / region_width
  )
}

# mod_dt: modification fraction per region (m6A and 5mC rows)
# cpg_dt: CpG block density per region
# random_mod_dt / random_cpg_dt: optional random-region data to overlay
make_region_plot <- function(mod_dt, cpg_dt, plot_title,
                             random_mod_dt = NULL, random_cpg_dt = NULL) {
  mod_dt <- data.table::copy(mod_dt)
  cpg_dt <- data.table::copy(cpg_dt)

  if (!is.null(random_mod_dt) && !is.null(random_cpg_dt)) {
    mod_dt <- data.table::rbindlist(list(data.table::copy(random_mod_dt), mod_dt), use.names = TRUE, fill = TRUE)
    cpg_dt <- data.table::rbindlist(list(data.table::copy(random_cpg_dt), cpg_dt), use.names = TRUE, fill = TRUE)
  }

  mod_dt[, modality := data.table::fifelse(modbase_code == "a", "m6A", "5mC")]

  ccre_classes <- setdiff(unique(cpg_dt$region_class), "random")
  class_levels <- c("random", ccre_classes)

  mod_dt[, region_class  := factor(region_class, levels = class_levels)]
  mod_dt[, modality      := factor(modality,     levels = c("m6A", "5mC"))]
  cpg_dt[, region_class  := factor(region_class, levels = class_levels)]
  cpg_dt[, region_source := factor(region_source, levels = c("random", "cCRE"))]

  mod_plot <- ggplot2::ggplot(mod_dt, ggplot2::aes(x = region_class, y = fraction_modified, fill = modality)) +
    ggplot2::geom_boxplot(position = ggplot2::position_dodge(width = 0.7), outlier.shape = NA, alpha = 0.45) +
    ggplot2::geom_point(position = ggplot2::position_jitterdodge(jitter.width = 0.12, dodge.width = 0.7), size = 2, alpha = 0.8) +
    ggplot2::scale_fill_manual(values = c("m6A" = "#7B3294", "5mC" = "#8C510A")) +
    ggplot2::coord_cartesian(ylim = c(0, 1)) +
    ggplot2::labs(title = paste0(plot_title, " - modification fractions"), x = "Region class", y = "Fraction modified", fill = "Modality") +
    ggplot2::theme_bw(base_size = 12)

  cpg_plot <- ggplot2::ggplot(cpg_dt, ggplot2::aes(x = region_class, y = cpg_feature_blocks_per_kb, fill = region_source)) +
    ggplot2::geom_boxplot(position = ggplot2::position_dodge(width = 0.7), outlier.shape = NA, alpha = 0.45) +
    ggplot2::geom_point(position = ggplot2::position_jitterdodge(jitter.width = 0.12, dodge.width = 0.7), size = 2, alpha = 0.8) +
    ggplot2::scale_fill_manual(values = c("random" = "#4D4D4D", "cCRE" = "#1B9E77")) +
    ggplot2::labs(title = paste0(plot_title, " - CpG block density"), x = "Region class", y = "CpG feature blocks per kb", fill = "Region source") +
    ggplot2::theme_bw(base_size = 12)

  mod_plot / cpg_plot + patchwork::plot_layout(heights = c(1, 1))
}

sample_nonoverlap_windows <- function(n_regions, chr_len, width_bp, exclude_gr) {
  starts0      <- integer()
  attempts     <- 0L
  max_attempts <- n_regions * 500L
  while (length(starts0) < n_regions && attempts < max_attempts) {
    attempts         <- attempts + 1L
    candidate_start0 <- sample.int(chr_len - width_bp + 1L, 1L) - 1L
    candidate_gr     <- GenomicRanges::GRanges(
      seqnames = default_chrom,
      ranges   = IRanges::IRanges(candidate_start0 + 1L, candidate_start0 + width_bp),
      strand   = "*"
    )
    if (!length(GenomicRanges::findOverlaps(candidate_gr, exclude_gr, ignore.strand = TRUE))) {
      starts0 <- c(starts0, candidate_start0)
      exclude_gr <- c(exclude_gr, candidate_gr)
    }
  }
  if (length(starts0) < n_regions) stop("Could not sample enough random non-cCRE, non-promoter windows.", call. = FALSE)
  starts0
}

# --- Analysis ---

bam_seqinfo   <- read_bam_seqinfo(bam_path)
tss_objects   <- read_gencode_tss(tss_path, chrom = default_chrom, window_bp = window_bp, seqinfo = bam_seqinfo)
promoter_gr   <- tss_objects$promoter_windows
cpg_bed12_all <- read_fiber_bed12(cpg_ft_path)

# All ENCODE Registry V4 cCRE classes:
#   PLS          = promoter-like signature
#   pELS         = proximal enhancer-like signature
#   dELS         = distal enhancer-like signature
#   CA-TF        = chromatin accessible with TF footprint
#   CA-CTCF      = chromatin accessible with CTCF binding
#   CA-H3K4me3   = chromatin accessible with H3K4me3
#   CA           = chromatin accessible only
#   TF           = TF binding only
#   Enhancers    = synthetic class: pELS + dELS combined
ccre_classes_base <- c("PLS", "pELS", "dELS", "CA-TF", "CA-CTCF", "CA-H3K4me3", "CA", "TF")

ccre_table_raw <- data.table::fread(
  ccre_path,
  header    = FALSE,
  col.names = c("chromosome", "start0", "end0", "ccre_id", "ccre_element_id", "ccre_class")
)[chromosome == default_chrom & ccre_class %in% ccre_classes_base]

# Synthetic "Enhancers" class = pELS + dELS rows relabelled
enhancer_rows <- data.table::copy(ccre_table_raw[ccre_class %in% c("pELS", "dELS")])
enhancer_rows[, ccre_class := "Enhancers"]

ccre_table_all <- data.table::rbindlist(
  list(ccre_table_raw, enhancer_rows),
  use.names = TRUE, fill = TRUE
)

# Use only raw (non-synthetic) rows for the exclusion GRanges
ccre_gr <- GenomicRanges::GRanges(
  seqnames = ccre_table_raw$chromosome,
  ranges   = IRanges::IRanges(ccre_table_raw$start0 + 1L, ccre_table_raw$end0),
  strand   = "*"
)

chr_len <- as.integer(GenomeInfoDb::seqlengths(bam_seqinfo)[default_chrom])
if (is.na(chr_len) || chr_len < window_bp) {
  stop(sprintf("Chromosome length for %s is not usable for random sampling.", default_chrom), call. = FALSE)
}

# Sample 5 non-overlapping random windows outside all promoters and cCREs
exclude_gr     <- c(promoter_gr, ccre_gr)
random_starts0 <- sample_nonoverlap_windows(5L, chr_len, window_bp, exclude_gr)
random_table   <- data.table::data.table(
  region_source = "random",
  region_class  = "random",
  chromosome    = default_chrom,
  start0        = random_starts0,
  end0          = random_starts0 + window_bp
)
random_table[, region_label := sprintf("random_%d|%s:%d-%d", .I, chromosome, start0 + 1L, end0)]

# Sample up to 5 regions per class (including synthetic Enhancers)
ccre_table <- ccre_table_all[, .SD[sample.int(.N, min(.N, 5L))], by = ccre_class]
ccre_table[, `:=`(
  region_source = "cCRE",
  region_class  = ccre_class,
  region_label  = sprintf("%s_%s|%s:%d-%d", ccre_class, ccre_id, chromosome, start0 + 1L, end0)
)]

random_cpg_summary <- data.table::rbindlist(
  lapply(seq_len(nrow(random_table)), function(i) summarize_region_cpg(random_table[i], cpg_bed12_all)),
  use.names = TRUE, fill = TRUE
)
random_mod_summary <- data.table::rbindlist(
  lapply(seq_len(nrow(random_table)), function(i) {
    rr <- random_table[i]
    data.table::rbindlist(list(
      summarize_region_modbam(rr, bamfiles, default_sample, bam_seqinfo, bpparam_modbam, "a"),
      summarize_region_modbam(rr, bamfiles, default_sample, bam_seqinfo, bpparam_modbam, "m")
    ), use.names = TRUE, fill = TRUE)
  }),
  use.names = TRUE, fill = TRUE
)

ccre_cpg_summary <- data.table::rbindlist(
  lapply(seq_len(nrow(ccre_table)), function(i) summarize_region_cpg(ccre_table[i], cpg_bed12_all)),
  use.names = TRUE, fill = TRUE
)
ccre_mod_summary <- data.table::rbindlist(
  lapply(seq_len(nrow(ccre_table)), function(i) {
    rr <- ccre_table[i]
    data.table::rbindlist(list(
      summarize_region_modbam(rr, bamfiles, default_sample, bam_seqinfo, bpparam_modbam, "a"),
      summarize_region_modbam(rr, bamfiles, default_sample, bam_seqinfo, bpparam_modbam, "m")
    ), use.names = TRUE, fill = TRUE)
  }),
  use.names = TRUE, fill = TRUE
)

# --- Plots ---

random_plot <- make_region_plot(
  random_mod_summary, random_cpg_summary,
  "Random non-promoter regions"
)

ccre_plot <- make_region_plot(
  ccre_mod_summary, ccre_cpg_summary,
  "cCRE regions (all classes) vs random",
  random_mod_dt = random_mod_summary,
  random_cpg_dt = random_cpg_summary
)

# --- Save outputs ---

write_gz_tsv(
  data.table::rbindlist(
    list(
      data.table::copy(random_cpg_summary)[, row_type := "cpg"],
      data.table::copy(random_mod_summary)[, row_type := "modbam"],
      data.table::copy(ccre_cpg_summary)[,  row_type := "cpg"],
      data.table::copy(ccre_mod_summary)[,  row_type := "modbam"]
    ),
    use.names = TRUE, fill = TRUE
  ),
  file.path(ccre_output_dir, sprintf("%s_%s_random_and_ccre_region_summary.tsv.gz", default_sample, default_chrom))
)

ggplot2::ggsave(
  file.path(ccre_output_dir, sprintf("%s_%s_random_nonpromoter_mod_cpg.pdf", default_sample, default_chrom)),
  random_plot, width = 12, height = 10
)
ggplot2::ggsave(
  file.path(ccre_output_dir, sprintf("%s_%s_ccre_vs_random_mod_cpg.pdf", default_sample, default_chrom)),
  ccre_plot, width = 18, height = 10
)

# --- Display ---
random_plot
ccre_plot
```

## CpG-only 5mC fraction

The previous section computes the 5mC fraction across **all C positions** in the BAM MM tag. 
Here we restrict the 5mC denominator and numerator to CpG-context cytosines only
```{r cpg-only-5mc, fig.width=18, fig.height=10}


# ── Variables ─────────────────────────────────────────────────────────────────
default_sample     <- "AL10_bc2178_19130"
default_chrom      <- "chr1"
mod_prob_threshold <- 0.9

results_root_dir <- "/project/spott/cshan/fiber-seq/results/promoter_summary"
ccre_output_dir  <- file.path(results_root_dir, "ccre")

bam_path <- file.path(
  "/project/spott/1_Shared_projects/LCL_Fiber_seq/preprocess_final_merged_samples",
  sprintf("%s.5mC.6mA.aligned.phased.bam", default_sample)
)
pbcpg_bed_path <- file.path(
  "/project/spott/cshan/fiber-seq/results/fire_CpG",
  default_sample,
  sprintf("%s_CPG.combined.bed.gz", default_sample)
)
prior_summary_path <- file.path(
  ccre_output_dir,
  sprintf("%s_%s_random_and_ccre_region_summary.tsv.gz", default_sample, default_chrom)
)

# ── Helpers (all functions needed; duplicates helpers chunk for standalone use) ─
assert_file_exists <- function(path, label = "file") {
  if (!file.exists(path)) stop(sprintf("Missing %s: %s", label, path), call. = FALSE)
  invisible(path)
}

write_gz_tsv <- function(x, path) {
  data.table::fwrite(x, file = path, sep = "\t")
  invisible(path)
}

read_bam_seqinfo <- function(bam_path) {
  targets <- Rsamtools::scanBamHeader(bam_path)[[1]]$targets
  GenomeInfoDb::Seqinfo(seqnames = names(targets), seqlengths = as.numeric(targets))
}

region_granges <- function(region_row) {
  GenomicRanges::GRanges(
    seqnames = region_row$chromosome,
    ranges   = IRanges::IRanges(region_row$start0 + 1L, region_row$end0),
    strand   = "*"
  )
}

read_modbam_summary_positions <- function(bamfiles, sample_name, promoter_windows,
                                          modbase_code, seqinfo, bpparam,
                                          modProbThreshold = 0.9) {
  se_summary <- footprintR::readModBam(
    bamfiles         = bamfiles,
    regions          = promoter_windows,
    modbase          = setNames(modbase_code, sample_name),
    level            = "summary",
    modProbThreshold = modProbThreshold,
    seqinfo          = seqinfo,
    trim             = TRUE,
    BPPARAM          = bpparam,
    verbose          = FALSE
  )
  gr             <- SummarizedExperiment::rowRanges(se_summary)
  total_calls    <- as.integer(SummarizedExperiment::assay(se_summary, "Nvalid")[, sample_name])
  modified_calls <- as.integer(SummarizedExperiment::assay(se_summary, "Nmod")[,  sample_name])
  data.table::data.table(
    call_chromosome  = as.character(GenomicRanges::seqnames(gr)),
    position         = as.integer(GenomicRanges::start(gr)),
    call_strand      = as.character(GenomicRanges::strand(gr)),
    sample_name      = sample_name,
    modbase_code     = modbase_code,
    total_calls      = total_calls,
    modified_calls   = modified_calls,
    unmodified_calls = total_calls - modified_calls,
    frac_modified    = data.table::fifelse(total_calls > 0,
                         modified_calls / total_calls, NA_real_)
  )[total_calls > 0]
}

make_region_plot <- function(mod_dt, cpg_dt, plot_title,
                             random_mod_dt = NULL, random_cpg_dt = NULL) {
  mod_dt <- data.table::copy(mod_dt)
  cpg_dt <- data.table::copy(cpg_dt)
  if (!is.null(random_mod_dt) && !is.null(random_cpg_dt)) {
    mod_dt <- data.table::rbindlist(list(data.table::copy(random_mod_dt), mod_dt),
                                    use.names = TRUE, fill = TRUE)
    cpg_dt <- data.table::rbindlist(list(data.table::copy(random_cpg_dt), cpg_dt),
                                    use.names = TRUE, fill = TRUE)
  }
  mod_dt[, modality := data.table::fifelse(modbase_code == "a", "m6A", "5mC")]
  ccre_classes <- setdiff(unique(cpg_dt$region_class), "random")
  class_levels <- c("random", ccre_classes)
  mod_dt[, region_class  := factor(region_class, levels = class_levels)]
  mod_dt[, modality      := factor(modality,     levels = c("m6A", "5mC"))]
  cpg_dt[, region_class  := factor(region_class, levels = class_levels)]
  cpg_dt[, region_source := factor(region_source, levels = c("random", "cCRE"))]
  mod_plot <- ggplot2::ggplot(mod_dt,
      ggplot2::aes(x = region_class, y = fraction_modified, fill = modality)) +
    ggplot2::geom_boxplot(position = ggplot2::position_dodge(width = 0.7),
                          outlier.shape = NA, alpha = 0.45) +
    ggplot2::geom_point(position = ggplot2::position_jitterdodge(
                          jitter.width = 0.12, dodge.width = 0.7), size = 2, alpha = 0.8) +
    ggplot2::scale_fill_manual(values = c("m6A" = "#7B3294", "5mC" = "#8C510A")) +
    ggplot2::coord_cartesian(ylim = c(0, 1)) +
    ggplot2::labs(title = paste0(plot_title, " - modification fractions"),
                  x = "Region class", y = "Fraction modified", fill = "Modality") +
    ggplot2::theme_bw(base_size = 12)
  cpg_plot <- ggplot2::ggplot(cpg_dt,
      ggplot2::aes(x = region_class, y = cpg_feature_blocks_per_kb, fill = region_source)) +
    ggplot2::geom_boxplot(position = ggplot2::position_dodge(width = 0.7),
                          outlier.shape = NA, alpha = 0.45) +
    ggplot2::geom_point(position = ggplot2::position_jitterdodge(
                          jitter.width = 0.12, dodge.width = 0.7), size = 2, alpha = 0.8) +
    ggplot2::scale_fill_manual(values = c("random" = "#4D4D4D", "cCRE" = "#1B9E77")) +
    ggplot2::labs(title = paste0(plot_title, " - CpG block density"),
                  x = "Region class", y = "CpG feature blocks per kb", fill = "Region source") +
    ggplot2::theme_bw(base_size = 12)
  mod_plot / cpg_plot + patchwork::plot_layout(heights = c(1, 1))
}

# ── Load saved output from random-region-cCRE chunk ──────────────────────────
assert_file_exists(prior_summary_path, "prior-chunk summary TSV")
prior_summary <- data.table::fread(prior_summary_path)

# Reconstruct region tables (one row per region, deduped from modbam rows)
modbam_rows  <- prior_summary[row_type == "modbam"]
random_table <- unique(modbam_rows[region_source == "random",
  .(region_source, region_class, region_label, chromosome,
    start0 = start - 1L, end0 = end)])
ccre_table   <- unique(modbam_rows[region_source == "cCRE",
  .(region_source, region_class, region_label, chromosome,
    start0 = start - 1L, end0 = end)])

# CpG density summaries for the plot's lower panel (unchanged from prior chunk)
random_cpg_summary <- prior_summary[row_type == "cpg" & region_source == "random"]
ccre_cpg_summary   <- prior_summary[row_type == "cpg" & region_source == "cCRE"]

# ── Re-read file-based inputs ─────────────────────────────────────────────────
assert_file_exists(bam_path,       "BAM")
assert_file_exists(pbcpg_bed_path, "pb-CpG-tools combined BED")

bamfiles       <- setNames(bam_path, default_sample)
bam_seqinfo    <- read_bam_seqinfo(bam_path)
bpparam_modbam <- BiocParallel::SerialParam()

# Read pb-CpG-tools BED: skip ##/# header lines, keep only chr of interest and
# type == "Total" (combined across haplotypes).
# Columns: chrom, begin (0-based), end, mod_score (0-100), type, cov,
#          est_mod_count, est_unmod_count, discretized_mod_score
pbcpg_bed <- data.table::fread(
  cmd = sprintf("zcat '%s' | grep -v '^#'", pbcpg_bed_path),
  col.names = c("chrom","begin","end","mod_score","type","cov",
                "est_mod_count","est_unmod_count","discretized_mod_score")
)[chrom == default_chrom & type == "Total"]

message(sprintf("pb-CpG BED loaded: %d CpG sites on %s", nrow(pbcpg_bed), default_chrom))

# ── m6A summarizer (BAM-based, unchanged) ────────────────────────────────────
summarize_region_modbam_m6a <- function(region_row, bamfiles, sample_name, seqinfo, bpparam) {
  position_dt <- read_modbam_summary_positions(
    bamfiles         = bamfiles,
    sample_name      = sample_name,
    promoter_windows = region_granges(region_row),
    modbase_code     = "a",
    seqinfo          = seqinfo,
    bpparam          = bpparam,
    modProbThreshold = mod_prob_threshold
  )
  total_calls    <- sum(position_dt$total_calls)
  modified_calls <- sum(position_dt$modified_calls)
  data.table::data.table(
    region_source           = region_row$region_source,
    region_class            = region_row$region_class,
    region_label            = region_row$region_label,
    chromosome              = region_row$chromosome,
    start                   = region_row$start0 + 1L,
    end                     = region_row$end0,
    region_width_bp         = region_row$end0 - region_row$start0,
    modbase_code            = "a",
    modified_calls          = modified_calls,
    unmodified_calls        = total_calls - modified_calls,
    total_calls             = total_calls,
    positions_with_coverage = nrow(position_dt),
    fraction_modified       = data.table::fifelse(
      total_calls > 0L, modified_calls / total_calls, NA_real_
    )
  )
}

# ── 5mC summarizer (pb-CpG-tools BED-based) ──────────────────────────────────
# For each region, sum est_mod_count and est_unmod_count over all CpG sites
# whose 0-based interval [begin, end) falls within [start0, end0).
# fraction_modified = est_mod_count / (est_mod_count + est_unmod_count).
summarize_region_pbcpg <- function(region_row, pbcpg_bed) {
  hits <- pbcpg_bed[
    chrom == region_row$chromosome &
    begin >= region_row$start0 &
    end   <= region_row$end0
  ]
  modified_calls   <- as.integer(round(sum(hits$est_mod_count)))
  unmodified_calls <- as.integer(round(sum(hits$est_unmod_count)))
  total_calls      <- modified_calls + unmodified_calls
  data.table::data.table(
    region_source           = region_row$region_source,
    region_class            = region_row$region_class,
    region_label            = region_row$region_label,
    chromosome              = region_row$chromosome,
    start                   = region_row$start0 + 1L,
    end                     = region_row$end0,
    region_width_bp         = region_row$end0 - region_row$start0,
    modbase_code            = "m",
    modified_calls          = modified_calls,
    unmodified_calls        = unmodified_calls,
    total_calls             = total_calls,
    positions_with_coverage = nrow(hits),
    fraction_modified       = data.table::fifelse(
      total_calls > 0L, modified_calls / total_calls, NA_real_
    )
  )
}

# ── Summarize: random regions ─────────────────────────────────────────────────
random_mod_cpgonly <- data.table::rbindlist(
  lapply(seq_len(nrow(random_table)), function(i) {
    rr <- random_table[i]
    data.table::rbindlist(list(
      summarize_region_modbam_m6a(rr, bamfiles, default_sample, bam_seqinfo, bpparam_modbam),
      summarize_region_pbcpg(rr, pbcpg_bed)
    ), use.names = TRUE, fill = TRUE)
  }),
  use.names = TRUE, fill = TRUE
)

# ── Summarize: cCRE regions ───────────────────────────────────────────────────
ccre_mod_cpgonly <- data.table::rbindlist(
  lapply(seq_len(nrow(ccre_table)), function(i) {
    rr <- ccre_table[i]
    data.table::rbindlist(list(
      summarize_region_modbam_m6a(rr, bamfiles, default_sample, bam_seqinfo, bpparam_modbam),
      summarize_region_pbcpg(rr, pbcpg_bed)
    ), use.names = TRUE, fill = TRUE)
  }),
  use.names = TRUE, fill = TRUE
)

# ── Plot ──────────────────────────────────────────────────────────────────────
cpgonly_plot <- make_region_plot(
  ccre_mod_cpgonly, ccre_cpg_summary,
  "cCRE regions - CpG-only 5mC vs random",
  random_mod_dt = random_mod_cpgonly,
  random_cpg_dt = random_cpg_summary
)

# ── Save ──────────────────────────────────────────────────────────────────────
write_gz_tsv(
  data.table::rbindlist(
    list(
      data.table::copy(random_mod_cpgonly)[, row_type := "modbam_cpgonly"],
      data.table::copy(ccre_mod_cpgonly)[,  row_type := "modbam_cpgonly"]
    ),
    use.names = TRUE, fill = TRUE
  ),
  file.path(ccre_output_dir,
    sprintf("%s_%s_random_and_ccre_cpgonly_5mc_summary.tsv.gz", default_sample, default_chrom))
)

ggplot2::ggsave(
  file.path(ccre_output_dir,
    sprintf("%s_%s_ccre_vs_random_cpgonly_5mc.pdf", default_sample, default_chrom)),
  cpgonly_plot, width = 18, height = 10
)

cpgonly_plot
```

## All-cCRE CpG-only 5mC (pb-CpG-tools, no subsampling)

Uses all chr1 cCREs from the ENCODE Registry V4. Methylation is computed entirely
from the pb-CpG-tools combined BED — no BAM query, no m6A.

```{r all-ccre-pbcpg-5mc, fig.width=14, fig.height=6}

# ── Variables ─────────────────────────────────────────────────────────────────
default_sample  <- "AL10_bc2178_19130"
default_chrom   <- "chr1"

results_root_dir <- "/project/spott/cshan/fiber-seq/results/promoter_summary"
ccre_output_dir  <- file.path(results_root_dir, "CpG_methylation_cCRE")

pbcpg_bed_path <- file.path(
  "/project/spott/cshan/fiber-seq/results/fire_CpG",
  default_sample,
  sprintf("%s_CPG.combined.bed.gz", default_sample)
)
ccre_path <- "/project/spott/cshan/annotations/GRCh38-cCREs.bed"

# ── Helpers ───────────────────────────────────────────────────────────────────
assert_file_exists <- function(path, label = "file") {
  if (!file.exists(path)) stop(sprintf("Missing %s: %s", label, path), call. = FALSE)
  invisible(path)
}

write_gz_tsv <- function(x, path) {
  data.table::fwrite(x, file = path, sep = "\t")
  invisible(path)
}

summarize_region_pbcpg_5mc <- function(region_row, pbcpg_bed) {
  hits <- pbcpg_bed[
    chrom == region_row$chromosome &
    begin >= region_row$start0 &
    end   <= region_row$end0
  ]
  modified_calls   <- as.integer(round(sum(hits$est_mod_count)))
  unmodified_calls <- as.integer(round(sum(hits$est_unmod_count)))
  total_calls      <- modified_calls + unmodified_calls
  data.table::data.table(
    region_class            = region_row$region_class,
    chromosome              = region_row$chromosome,
    start                   = region_row$start0 + 1L,
    end                     = region_row$end0,
    region_width_bp         = region_row$end0 - region_row$start0,
    n_cpg_sites             = nrow(hits),
    modified_calls          = modified_calls,
    unmodified_calls        = unmodified_calls,
    total_calls             = total_calls,
    fraction_modified       = data.table::fifelse(
      total_calls > 0L, modified_calls / total_calls, NA_real_
    )
  )
}

# ── Load data ─────────────────────────────────────────────────────────────────
assert_file_exists(pbcpg_bed_path, "pb-CpG-tools combined BED")
assert_file_exists(ccre_path,      "cCRE BED")

pbcpg_bed <- data.table::fread(
  cmd = sprintf("zcat '%s' | grep -v '^#'", pbcpg_bed_path),
  col.names = c("chrom", "begin", "end", "mod_score", "type", "cov",
                "est_mod_count", "est_unmod_count", "discretized_mod_score")
)[chrom == default_chrom & type == "Total"]

message(sprintf("pb-CpG BED: %d CpG sites on %s", nrow(pbcpg_bed), default_chrom))

ccre_classes_base <- c("PLS", "pELS", "dELS", "CA-TF", "CA-CTCF", "CA-H3K4me3", "CA", "TF")

ccre_table_raw <- data.table::fread(
  ccre_path, header = FALSE,
  col.names = c("chromosome", "start0", "end0", "ccre_id", "ccre_element_id", "ccre_class")
)[chromosome == default_chrom & ccre_class %in% ccre_classes_base]

# Synthetic Enhancers class = pELS + dELS combined
enhancer_rows <- data.table::copy(ccre_table_raw[ccre_class %in% c("pELS", "dELS")])
enhancer_rows[, ccre_class := "Enhancers"]

ccre_table_all <- data.table::rbindlist(
  list(ccre_table_raw, enhancer_rows),
  use.names = TRUE, fill = TRUE
)
ccre_table_all[, region_class := ccre_class]

message(sprintf("cCREs loaded: %d regions across %d classes on %s",
  nrow(ccre_table_raw), length(ccre_classes_base), default_chrom))

# ── Summarize all cCREs ───────────────────────────────────────────────────────
ccre_5mc_all <- data.table::rbindlist(
  lapply(seq_len(nrow(ccre_table_all)), function(i) {
    summarize_region_pbcpg_5mc(ccre_table_all[i], pbcpg_bed)
  }),
  use.names = TRUE, fill = TRUE
)

# Order classes: base classes in defined order, then synthetic Enhancers
class_order <- c(ccre_classes_base, "Enhancers")
ccre_5mc_all[, region_class := factor(region_class, levels = class_order)]

message(sprintf("Regions with CpG coverage: %d / %d",
  sum(ccre_5mc_all$total_calls > 0L, na.rm = TRUE), nrow(ccre_5mc_all)))

# ── Plot ──────────────────────────────────────────────────────────────────────
n_per_class <- ccre_5mc_all[!is.na(fraction_modified), .N, by = region_class]
x_labels    <- setNames(
  sprintf("%s\n(n=%d)", n_per_class$region_class, n_per_class$N),
  n_per_class$region_class
)

all_ccre_5mc_plot <- ggplot2::ggplot(
    ccre_5mc_all[!is.na(fraction_modified)],
    ggplot2::aes(x = region_class, y = fraction_modified)
  ) +
  ggplot2::geom_boxplot(fill = "#8C510A", alpha = 0.45, outlier.shape = NA) +
  ggplot2::scale_x_discrete(labels = x_labels) +
  ggplot2::scale_y_continuous(limits = c(0, 1), expand = ggplot2::expansion(mult = 0.02)) +
  ggplot2::labs(
    title = sprintf("CpG-only 5mC fraction by cCRE class — all %s cCREs (%s)",
                    default_chrom, default_sample),
    x     = "cCRE class",
    y     = "CpG methylation fraction (pb-CpG-tools)"
  ) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(axis.text.x = ggplot2::element_text(size = 10))

# ── Save ──────────────────────────────────────────────────────────────────────
write_gz_tsv(
  ccre_5mc_all,
  file.path("/project/spott/cshan/fiber-seq/results/promoter_summary/CpG_methylation_cCRE/",
    sprintf("%s_%s_all_ccre_pbcpg_5mc.tsv.gz", default_sample, default_chrom))
)

ggplot2::ggsave(
  file.path(ccre_output_dir,
    sprintf("%s_%s_all_ccre_pbcpg_5mc.pdf", default_sample, default_chrom)),
  all_ccre_5mc_plot, width = 14, height = 6
)

saveRDS(ccre_5mc_all, file = "/project/spott/cshan/fiber-seq/results/promoter_summary/CpG_methylation_cCRE/ccre_5mc_all.rds")
all_ccre_5mc_plot
```

## All-cCRE CpG-only 5mC metaprofile (aggregate signal around cCRE centers)

Average CpG methylation fraction binned by distance from each cCRE center.
Uses all chr1 cCREs from the previous chunk (`pbcpg_bed`, `ccre_table_all`, `class_order`).

```{r all-ccre-pbcpg-5mc-metaprofile, fig.width=16, fig.height=10}
# ccre_5mc_all <- fread("/project/spott/cshan/fiber-seq/results/promoter_summary/CpG_methylation_cCRE/AL10_bc2178_19130_chr1_all_ccre_pbcpg_5mc_metaprofile.tsv.gz")
# class_order <- c(ccre_classes_base, "Enhancers")
# ccre_5mc_all[, region_class := factor(region_class, levels = class_order)]

message(sprintf("Regions with CpG coverage: %d / %d",
  sum(ccre_5mc_all$total_calls > 0L, na.rm = TRUE), nrow(ccre_5mc_all)))

# ── Parameters ────────────────────────────────────────────────────────────────
meta_window_bp <- 1000L   # flanking bp each side of cCRE center
bin_size       <- 20L     # bp per bin

# ── Build windows centred on each cCRE ───────────────────────────────────────
ccre_wins <- data.table::copy(ccre_table_all)[, .(
  chrom      = chromosome,
  region_class,
  center     = as.integer((start0 + end0) / 2L),
  win_start  = as.integer((start0 + end0) / 2L) - meta_window_bp,
  win_end    = as.integer((start0 + end0) / 2L) + meta_window_bp
)]
ccre_wins[win_start < 0L, win_start := 0L]   # clamp to chromosome start

# ── Prepare CpG table for interval overlap ────────────────────────────────────
# pbcpg_bed already filtered to default_chrom & type == "Total"
cpg_for_join <- pbcpg_bed[, .(
  chrom           = chrom,
  cpg_pos         = begin,           # 0-based; treat as point interval
  cpg_pos_end     = begin,
  est_mod_count   = as.numeric(est_mod_count),
  est_unmod_count = as.numeric(est_unmod_count)
)]
data.table::setkey(cpg_for_join, chrom, cpg_pos, cpg_pos_end)
data.table::setkey(ccre_wins,    chrom, win_start, win_end)

# foverlaps: for each CpG point, find all cCRE windows it falls in
overlaps <- data.table::foverlaps(
  cpg_for_join, ccre_wins,
  by.x    = c("chrom", "cpg_pos", "cpg_pos_end"),
  by.y    = c("chrom", "win_start", "win_end"),
  type    = "within",
  nomatch = NULL
)

overlaps[, rel_pos := cpg_pos - center]
overlaps[, bin     := as.integer(round(rel_pos / bin_size) * bin_size)]

# ── Aggregate per (class, bin) ────────────────────────────────────────────────
meta_dt <- overlaps[, .(
  modified_calls   = sum(est_mod_count),
  unmodified_calls = sum(est_unmod_count),
  total_calls      = sum(est_mod_count + est_unmod_count),
  n_cpg_obs        = .N
), by = .(region_class, bin)]

meta_dt[, fraction_modified := data.table::fifelse(
  total_calls > 0, modified_calls / total_calls, NA_real_
)]

# Fill in missing bins so lines are continuous
all_bins  <- seq(-meta_window_bp, meta_window_bp, by = bin_size)
meta_full <- data.table::CJ(
  region_class = unique(ccre_table_all$region_class),
  bin          = all_bins
)
meta_full <- meta_dt[meta_full, on = .(region_class, bin)]
meta_full[, region_class := factor(region_class, levels = class_order)]

# Count of cCREs per class (for facet labels)
n_ccre_class <- ccre_table_all[, .N, by = .(region_class = region_class)]
facet_labels <- setNames(
  sprintf("%s\n(n=%d cCREs)", n_ccre_class$region_class, n_ccre_class$N),
  n_ccre_class$region_class
)

# ── Plot ──────────────────────────────────────────────────────────────────────
meta_plot <- ggplot2::ggplot(
    meta_full[!is.na(fraction_modified)],
    ggplot2::aes(x = bin, y = fraction_modified)
  ) +
  ggplot2::geom_line(color = "#8C510A", linewidth = 0.4, na.rm = TRUE) +
  ggplot2::geom_smooth(
    method    = "loess",
    span      = 0.15,
    se        = FALSE,
    color     = "#8C510A",
    linewidth = 0.9,
    na.rm     = TRUE
  ) +
  ggplot2::geom_vline(
    xintercept = 0,
    linetype   = "dashed",
    color      = "grey40",
    linewidth  = 0.4
  ) +
  ggplot2::scale_x_continuous(
    breaks = seq(-meta_window_bp, meta_window_bp, 200)
  ) +
  ggplot2::scale_y_continuous(
    expand = ggplot2::expansion(mult = 0.02)
  ) +
  ggplot2::facet_wrap(
    ~region_class,
    nrow     = 3,
    scales   = "free_y",
    labeller = ggplot2::as_labeller(facet_labels)
  ) +
  ggplot2::labs(
    title = sprintf(
      "CpG-only 5mC metaprofile around cCRE centers — all %s cCREs (%s)",
      default_chrom, default_sample
    ),
    x = "Distance from cCRE center",
    y = "CpG methylation fraction"
  ) +
  ggplot2::theme_bw(base_size = 11) +
  ggplot2::theme(
    strip.text  = ggplot2::element_text(size = 9),
    axis.text.x = ggplot2::element_text(size = 8)
  )

# ── Save ──────────────────────────────────────────────────────────────────────
# write_gz_tsv(
#   meta_full,
#   file.path(ccre_output_dir,
#     sprintf("%s_%s_all_ccre_pbcpg_5mc_metaprofile.tsv.gz", default_sample, default_chrom))
# )



ggplot2::ggsave(
  file.path(ccre_output_dir,
    sprintf("100bp_window_20bp_bin%s_%s_all_ccre_pbcpg_5mc_metaprofile.pdf", default_sample, default_chrom)),
  meta_plot, width = 16, height = 10
)

meta_plot
```
